map3k1 is required for spatial restriction of progenitor differentiation in planarians

Version 1

Abstract

Planarian regeneration and tissue turnover involve fate specification in pluripotent stem cells called neoblasts. Neoblasts select fates through the expression of fate-specific transcription factors (FSTFs), generating specialized neoblasts. Specialized neoblasts are spatially intermingled and can be dispersed broadly, frequently being far from their target tissue. The post-mitotic progeny of neoblasts, serving as progenitors, migrate and differentiate into mature cell types. Pattern formation is thus strongly influenced by the migratory assortment and differentiation of fate-specified progenitors in precise locations, which we refer to as progenitor targeting. This central step of pattern maintenance and formation, however, is poorly understood. Here, we describe a requirement for the conserved map3k1 gene in targeting, restricting post-mitotic progenitor differentiation to precise locations. RNAi of map3k1 causes ectopic differentiation of eye progenitors along their migratory path, resulting in dispersed ectopic eyes and eye cells. Other neural tissues similarly display ectopic posterior differentiation and pharynx cells emerge dispersed laterally and anteriorly in map3k1 RNAi animals. Ectopic differentiated cells are also found within the incorrect organs after map3k1 RNAi, and ultimately teratomas form. These findings implicate map3k1 signaling in controlling the positional regulation of progenitor behavior – restricting progenitor differentiation to targeted locations in response to external cues in the local tissue environment.

Introduction

Planarians are capable of regenerating any missing body part through the action of pluripotent stem cells called neoblasts (Reddien, 2018). Neoblasts also maintain all cell types of the adult body through a process of constitutive cell turnover. Because planarians have >125 different adult cell types, these stem cells must be capable of choosing among a large array of possible cell fates (Fincher et al., 2018; Plass et al., 2018; Zeng et al., 2018; King et al., 2024). Fate specification can occur in neoblasts through the activation of transcription factors called fate-specific transcription factors (FSTFs), producing specialized neoblasts (Reddien, 2022). Specialized neoblasts divide and can produce daughter cells that act as migratory precursors (post-mitotic progenitors) for differentiated cell types (Eisenhoffer et al., 2008; van Wolfswinkel et al., 2014; Reddien, 2022; King et al., 2024).

Fate choice in neoblasts can be regulated by position. For instance, eye-specialized neoblasts are formed in roughly the anterior third of the animal (Lapan and Reddien, 2011, 2012). However, this spatial regulation of stem cell fate specification is coarse, when compared to the precise positioning of cell differentiation to form functional, complex tissue architecture (Lapan and Reddien, 2011, 2012; Adler et al., 2014; Scimone et al., 2014a; van Wolfswinkel et al., 2014; Park et al., 2023). Because specialized neoblasts are produced in broad regions, they are often found far from their target tissue. Indeed, specialized neoblasts can frequently be found closer to other differentiated cell types than to the differentiated tissue they are fated to mature into (Park et al., 2023). Furthermore, neoblasts are specified in a highly spatially intermingled manner, in heterogenous neoblast neighborhoods (Park et al., 2023). for instance, a muscle-specialized neoblast could have a neural, intestinal, epidermal, eye, or other specialized neoblast as its nearest neoblast neighbor (Park et al., 2023). These observations suggest that the regulation of differentiation programs for post-mitotic migratory progenitors is a crucial aspect to patterning and maintaining the various tissues in the animal. Neoblasts themselves are not highly migratory, however (Saló and Baguñà, 1985; Guedelhoefer and Sánchez Alvarado, 2012; Abnave et al., 2017). By contrast, the post-mitotic descendant cells of neoblasts can remain immature for a number of days – serving as progenitors – and can be migratory (Eisenhoffer et al., 2008; Wenemoser and Reddien, 2010; Guedelhoefer and Sánchez Alvarado, 2012; van Wolfswinkel et al., 2014; Abnave et al., 2017; Park et al., 2023). These findings generate a model for pattern maintenance during tissue turnover and formation in planarian regeneration in which specialized neoblasts generate post-mitotic progenitors that assort and target precise locations for differentiation through migration (Park et al., 2023).

From a messy state of planarian progenitor formation, order must arise. Terminal differentiation of post-mitotic progenitors must be regulated to occur at precise positions to prevent disordered differentiation along migratory trails. Defects in this process could lead to errors in tissue pattern, such as supernumerary organs, ectopic differentiation of single cells, incorrect incorporation of a particular cell type into the wrong tissue, and teratoma formation. Therefore, understanding how migratory progenitors know when and where to terminally differentiate into a mature stationary cell is a central problem for understanding how systems of migratory progenitors generate and maintain pattern. Little is known about the molecular regulation of differentiation, restricting it to occur in precise locations. Given that progenitor migratory targeting and differentiation are the predominant processes involved in planarian adult tissue patterning, identification of the molecular processes regulating these behaviors is central for understanding adult regeneration and pattern maintenance. Planarians present the opportunity to uncover general mechanisms underlying migratory progenitor targeting and differentiation regulation, which could apply to a myriad of developmental and regenerative contexts across the animal kingdom.

Results

map3k1 RNAi results in the ectopic posterior differentiation of eye cells along the AP axis

We sought to understand how the differentiation of fate-specified progenitors is spatially regulated to control planarian patterning through RNAi studies. The RNAi of a single gene (map3k1, dd_5198) encoding a MAP3K1-like signaling protein resulted in a striking phenotype of ectopic eye formation that was distinct in nature from previously described planarian patterning phenotypes, and was suggestive of having a candidate role in progenitor targeting (Figure 1A). We therefore selected this gene for detailed investigation given the robust and unique nature of its patterning phenotype. Planarian map3k1 encodes a member of the MAP3K1 subclass of MAP3K signaling proteins. MAP3K1 proteins, including the planarian ortholog, possess a unique domain architecture among MAP3Ks. Aside from its characteristic kinase domain, MAP3K1 has other unique domains: a PHD-like RING finger, a SWIM-type RING finger, and a TOG-like domain. These domains have been implicated in non-canonical MAP kinase signaling cascades that enable MAP3K1 to act as a ubiquitin ligase, a scaffold protein, and to signal through MAP kinase proteins (Figure 1- figure supplement 1).

map3k1 RNAi results in ectopic eyes and ectopic isolated eye cells.

A. Live images (upper panels) showing posterior eyes in map3k1 RNAi animals. Upper panels, 6 feedings of RNAi over 36 days. FISH (bottom panels): single optic cup and photoreceptor cells are present ectopically with RNA probes for opsin (photoreceptors), a catalase/tyrosinase/glut3 pool (optic cup), and an anti-Arrestin antibody (labeling photoreceptors and their projections). B. Examples showing a diversity of ectopic eyes and eye cells in map3k1 RNAi animals. A, B. Dorsal up; Scale bar, 200μm. C. Schematic comparing previously identified eye-patterning phenotypes with the map3k1 RNAi phenotype. D. FISH of single cells differentiated proximal to the pharynx and in the tail of map3k1 RNAi animals. Dorsal up; scale bar, 100μm. E. FISH examples of ovo+; smedwi-1+ cells in the tail. F. Left graph depicts the number of ectopic optic cup differentiation events per animal along the AP axis over 10, 16, and 26 days. Right graph depicts the number of ectopic photoreceptor neurons at 26 days of map3k1 RNAi; See also Figure 1- figure supplement 1.

map3k1 RNAi results in differentiated eye cells throughout the AP axis, and no overt change in eye progenitor distribution.

A. Domain structure of human and Schmidtea mediterranea MAP3K1. % identity/ similarity is displayed above each domain. B. FISH of map3k1 RNAi animals at 10 days, 16 days, and 6 weeks of RNAi shows optic cup (catalase1, tyrosinase, glut3) in the posterior at early timepoints. C. FISH with an RNA probe to the map3k1 gene in a wild-type animal. Scale bar, 200μm for full body images; Scale bar, 100μm for closeup of the brain. D. FISH showing double-positive ovo+ ; smedwi-1+ cells near the posterior half of the pharynx. Scale bars, 50μm. E. quantification of smedwi-1 high, smedwi-1 low, and smedwi-1 negative ovo+ cells in the tails of map3k1 RNAi and control animals. No significant change is observed between the number of smedwi-1+ ; ovo+ cells in the tail between map3k1 RNAi and control animals.

The bilaterally symmetric anterior location of planarian eyes represents the normal targeting location of eye progenitors (Lapan and Reddien, 2011, 2012). RNAi of map3k1 resulted in the gradual emergence of ectopic eyes posterior to the normal eye location (Figure 1A,B). Ectopic eyes displayed some variance in their posterior and medial-lateral (ML) positions in the pre-pharyngeal animal region (Figure 1A,B). Fluorescent in situ hybridization (FISH) experiments showed that both optic cup and photoreceptor cells were present ectopically (Figure 1A,B). Ectopic optic cup and photoreceptor cells were observed both as clusters of cells that nucleated and grew into additional eyes, as well as individual, isolated eye cells (Figure 1A,B; Figure 1- figure supplement 1B). Isolated, individual photoreceptor neuron or optic cup cell differentiation is rarely observed in the wild-type state. Surprisingly, although ectopic eye cells appeared most frequently in the normal zone of eye progenitor specification, they also appeared along the entire AP axis, including in the posterior half of the tail – far from the normal eye-progenitor specification zone (Figure 1C, Figure 1- figure supplement 1B). Most animals had at least one ectopic eye cell in the tail after 26 days of map3k1 RNAi. The map3k1 RNAi phenotype did not appear to result from a anterior-to-posterior progression of eye formation, because isolated eye cells were observed in RNAi animal tails even at early RNAi timepoints when no, or very few other ectopic differentiated eye cells appeared between the head and tail (Figure 1C, F, Figure 1- figure supplement 1B). The very posterior ectopic eye cells that appeared in map3k1 RNAi animals were more sparse than those in the head, and more likely to appear as singletons rather than aggregates (Figure 1C, Figure 1- figure supplement 1B).

The location where dispersed eye progenitors migrate towards and undergo terminal differentiation is referred to as the target zone (TZ) (Atabay et al., 2018). A variety of genes can be inhibited to result in eye formation in ectopic positions along the anterior-posterior (AP) axis: for instance, posteriorly for RNAi of nou darake (ndk) (Cebria et al., 2002) or wntA (Kobayashi et al., 2007), and anteriorly for RNAi of notum (Hill and Petersen, 2015) and nr4A (Li et al., 2019). Furthermore, wnt5 RNAi causes ectopic lateral eye formation and slit RNAi cause medial eye formation (Oderberg et al., 2017; Atabay et al., 2018). However, in all of these cases ectopic eyes appear along particular AP or ML trajectories (Figure 1D, summary cartoon), consistent with alteration of the eye progenitor TZ along a single (AP or ML) axis, but still with differentiation constrained to occur in a particular position on the orthogonal axis. By contrast, map3k1 RNAi ectopic eyes were more diffuse, often being ectopic in both AP and ML position. This distinct patterning phenotype raised the possibility that map3k1 was required for a patterning process not previously perturbed by inhibition of prior patterning genes.

Patterning in planarians prominently involves constitutive and regional expression of genes constituting positional information referred to as position control genes (PCGs) (Reddien, 2018). PCGs are predominantly expressed in planarian muscle (Witchley et al., 2013). slit, wnt5, ndk, nr4A, and notum are regulated in their spatial expression and are components of this PCG patterning system. Therefore, these genes likely influence progenitor-extrinsic cues that guide progenitor targeting. By contrast, map3k1 is not expressed overtly in a spatially restricted manner, instead being expressed broadly across tissues (Figure 1- figure supplement 1C).

The appearance of eye cells in the tail of map3k1 RNAi animals suggests that some eye progenitor specification is likely occurring in the tails of these animals. Eye specialized neoblasts can be recognized by the co-expression of the eye-specific FSTF ovo (Lapan and Reddien, 2012) and the neoblast marker smedwi-1 (Reddien et al., 2005). Indeed, ovo+; smedwi-1+ neoblasts were observed at low frequency in the midbody and tail of map3k1 RNAi animals (Figure 1E; Figure 1- figure supplement 1D, E). The frequency of these observed rare ovo+ neoblasts was not substantially different from a low frequency observed in control animals, raising the possibility that low frequency sporadic eye neoblast specification events in the body can lead to ectopic differentiation when map3k1 is inhibited.

map3k1 prevents the posteriorization of some but not all anterior cell types

To assess whether the ectopic differentiation of progenitors that occurred in map3k1 RNAi animals was specific to the eye, we visualized other features of differentiated tissue patterns. For example, lateral branches are normally associated specifically with the planarian brain (Hyman, 1951). However, after map3k1 RNAi, ectopic brain branches emerged from the two ventral nerve cords – most commonly observed in the anterior and mid-body, extending to the far posterior of the animal over long-term (>6 weeks) RNAi in uninjured animals (Figure 2A, Figure 2- figure supplement 1A). Ectopic eyes sent axons that traveled along branches and main nerve cord tracts (Figure 2A, Figure 2- figure supplement 1A). Ectopic brain branches that emerged from ventral nerve cords contained GluR+ neurons, confirming that they are at least partly comprised of normally brain-branch-restricted neurons (Figure 2B). GluR+ neurons are similarly found in ectopic brain branches observed after RNAi of ndk (Cebria et al., 2002).

map3k1 RNAi results in ectopic neuron classes and gland cells along the AP axis.

A. map3k1 RNAi animals exhibit posterior ectopic brain branches with chat+ neurons and gluR+ (dd_16476) neurons. Photoreceptor axons extend along ventral nerve cords and into ectopic brain branches. Ventral up; Scale bars, 200μm; Closeup images, scale bar, 100μm. B. FISH images showing no change in cintillo+ neuron distribution, but expansion of dd_17258+ cells down the entire AP axis. ventral up; scale bars, 200μm; See also Figure 2- figure supplement 1. C. Graph showing distribution of dd_17258+ cells along the AP axis in control and map3k1 RNAi animals. D. Quantification of cells in the tail for dd_7131 and dd_8476. More cells are observed in the tails of map3k1 RNAi animals compared to control animals; Student t-test using Graphpad Prism software ***p<.0001; **p<.001 . E. FISH examples of unaffected (dd_9223) and affected (dd_7131 and dd_8476) parenchymal cell types in map3k1 RNAi animals. ventral up; scale bar, 200μm. Numbers of animals are represented on the bottom left of each panel.

map3k1 RNAi results ectopic brain branches posterior to the brain and posterior expansion of some neural cell types.

A. Close-up FISH of ventral nerve cord (VNC) and branches in a map3k1 RNAi animal. Photoreceptor projections (anti-Arrestin) are travel along the VNC and into ectopic branches labeled by the pan-neural markers chat and pc2. Scale bar, 100μm. B. FISH of dd_17258+ cells and optic cup cells (catalase1, tyrosinase, glut3), together with antibody labeling of photoreceptors (anti-Arrestin), extending down the AP axis. All scale bars, 200μm, unless otherwise specified in the lower left. C. FISH showing expansion of dd_17258 cells down the entire AP axis in a map3k1 RNAi animal. Scale bar, 200μm. D. FISH showing an unchanged neural cell type (gad) in map3k1 RNAi and control animals. Scale bar, 200μm.

Another anterior neural population, marked with an RNA probe to dd_17258, also displayed ectopic differentiation across the AP axis of map3k1 RNAi animals – including through to the tip of the tail (Figure 2C, Figure 2- figure supplement 1B,C). By contrast, the AP location of cintillo+ neurons, which display posteriorization in ndk RNAi animals (Scimone et al., 2016), remained unaffected in map3k1 RNAi animals (Figure 2C). Another anterior-restricted neural population (glutamic acid decarboxylase, gad+) also failed to show posteriorization following map3k1 RNAi (Figure 2- figure supplement 1D). These findings suggest that map3k1 is required for the normal AP restriction of a subset of neural cell types during tissue turnover. The cell types that displayed posteriorization following map3k1 RNAi also maintained their general tissue structure in their normal locations, rather than being lost anteriorly and replaced with a more posterior domain.

To determine if tissue posteriorization occurred in non-neuronal cell types, we assayed particular gland cell populations that predominantly reside in the head, with a small fraction extending down to the pharynx and rarely to the tail. map3k1 RNAi animals showed posteriorization of gland cell distributions for the primarily anterior dd_7131+ and dd_8476+ populations (Figure 2D-F). As was the case with the neurons, not all gland cell types were affected by map3k1 RNAi. dd_9223+ cells, which are normally restricted to the anterior, did not change in distribution (Figure 2D). These data indicate that map3k1 broadly affects the spatial distribution of differentiated cell types for multiple tissues during tissue maintenance through turnover, but that its role in regulating differentiation is restricted to a subset of progenitor populations.

map3k1 inhibition causes ectopic anterior differentiation of pharynx progenitors

Another regional tissue that is maintained through turnover from regional progenitors is the planarian pharynx. FoxA is expressed in neoblasts surrounding the centrally located pharynx, specifying progenitors for this organ (Adler et al., 2014; Scimone et al., 2014a). These progenitors must enter the pharynx through a connection to the body at the anterior end of the organ. After map3k1 RNAi, vitrin+ pharynx cells appeared at the anterior domain of the predicted FoxA+ zone, through which pharynx progenitors are migrating, as well more anterior to the normal FoxA+ zone (Figure 3A). vitrin+ cells clustered together to form small structures containing pharyngeal cells between the original pharynx and the brain (Figure 3A). Most animals developed one main ectopic pharynx cell cluster anterior to the original pharynx, whereas others developed multiple anterior structures that could also exist lateral to the midline (Figure 3A). This variable placement of ectopic pharyngeal cells on the ML axis is reminiscent of the patterning defect observed with eyes described above. Interestingly, many progenitors in the FoxA zone appear to differentiate as singletons – single pharynx cells, a phenotype not previously observed with other patterning phenotypes, and also bearing similarity to the ectopic eye cell defect.

map3k1 RNAi results in pharynx cell types in ectopic anterior locations.

A. FISH images showing anterior expansion and disorganization of vitrin+ (pharynx-specific marker) cells along the AP-axis. Ventral up; scale bars, 100μm. B. FISH showing anterior expansion of NB.22.1e (mouth and esophagus) cells along the AP axis. map3k1 RNAi animals exhibit a low frequency of NB.22.1e cells in the tail, posterior to the normal mouth location. C. FISH shows anterior expansion of mhc1+ cells in map3k1 RNAi animals compared to control animals. ventral up, scale bar, 100μm. D. FISH shows anterior expansion of dd_554 cells (pharynx progenitors) in map3k1 RNAi animals compared to control animals. Ventral, up. Scale bar, 100μm. E. FISH shows FoxA+; smedwi-1+ cells close to the brain lobes and eyes. Scale bar, 50μm.

The planarian mouth is an epidermal opening associated with the posterior end of the pharynx. Ectopic anterior mouth cells marked by RNA transcripts for the gene NB.22.1e were also observed following map3k1 RNAi. Ectopic mouth cells appeared as an anterior streak of cells with NB.22.1e transcripts, originating from the original mouth. In addition, ectopic sparse, scattered mouth cells were also present, sometimes lateral to the midline (Figure 3B). Rarely (n=2/12 animals), an ectopic focus of NB.22.1e+ mouth cells was observed in the tail, posterior to the pharynx (Figure 3B). We also examined the pattern of cells expressing mhc1, which marks pharyngeal muscle. Similar to the case of vitrin+ cells, scattered ectopic foci of pharyngeal muscle cells were observed in the anterior half of the animal (Figure 3C).

Because differentiated pharynx cells were observed outside of the normal progenitor specification zone, we considered the possibility that pharynx progenitor specification itself occurred in ectopic locations following map3k1 RNAi. dd_554 transcripts mark a post-mitotic pharynx progenitor population (Zhu et al., 2015). Ectopic dd_554 cells were observed anterior to the pharynx in map3k1 RNAi animals, including in foci (Figure 3D). FoxA+; smedwi-1+ cells include prominently pharynx progenitors. A small frequency of animals exhibited FoxA+; smedwi-1+ cells between the lobes of the cephalic ganglia of long term map3k1 RNAi animals. This location is anterior to the normal progenitor specification zone for the pharynx (Figure 3E), which is typically centrally restricted on the AP axis (Adler et al., 2014; Scimone et al., 2014a). These findings are consistent with the interpretation that these FoxA+; smedwi-1+ cells possibly include pharynx progenitors and that stem cell fate specification may occur outside of the normally spatially restricted domains in map3K1 RNAi animals, at least in the case of the pharynx.

Complete posterior duplication of the pharynx has previously been observed with ptk7, ndl-3, and wntP-2 RNAi (Scimone et al., 2016; Hill and Petersen, 2018), anterior duplication has been observed after roboA RNAi (Cebria and Newmark, 2007), and lateral duplication has been observed after wnt5 RNAi (Gurley et al., 2010). The map3k1 RNAi phenotype differs from these other patterning phenotypes, in that it involves greater disorganization, small clusters of pharyngeal cells, and even single pharyngeal cells. This scenario bears similarity to the phenotype for the eye: disorganized ectopic tissue differentiation including in small clusters and single cells. These scenarios suggest defects in the regulation of progenitor targeting for differentiation in precise locations.

The targeting and maintenance of tissues after map3k1 RNAi

Post-mitotic progenitors normally migrate to their target zone (TZ) and differentiate at this location. Prior work indicates that in addition to the TZ, the target tissue itself can incorporate its fate-specified progenitors and promote progenitor differentiation, even in ectopic locations (outside of the TZ) (Atabay et al., 2018; Hill and Petersen, 2018). For instance, surgical manipulations (including transplantation) that lead to ectopic eyes outside of the TZ in wild-type animals result in stable maintenance of those ectopic eyes through progenitor incorporation and differentiation as part of turnover (Atabay et al., 2018). This is enabled by the fact that progenitors in wild-type animals are specified in broad regions, giving ectopic differentiated tissues access to a constant supply of progenitors (Lapan and Reddien, 2011, 2012). Thus, at least two components appear to be capable of promoting progenitor differentiation: the TZ and the target tissue. The pharynx shows similar properties to the case of the eye – with an ectopic pharynx being maintained through progenitor incorporation and differentiation (Hill and Petersen, 2018). A surgically transplanted ectopic organ (outside of its TZ), however, will not regenerate upon its removal. Because the original TZ location of the tissue is unchanged, progenitors will target the correct location after resection of an ectopic organ (Atabay et al., 2018; Hill and Petersen, 2018).

Given the above reasoning, ectopic eyes in map3k1 RNAi animals could in principle be explained by TZ movement or expansion, if map3k1 primarily acted to control TZ locations. To assess whether TZ alteration, or some other explanation for the map3k1 RNAi phenotype is more likely, we resected all visible eyes in map3k1 RNAi animals to determine the location of new progenitor differentiation in the absence of a target tissue. These eye-resected map3k1 RNAi animals grew back eyes at a comparable rate to control animals, and in the expected wild-type (normal TZ) location (Figure 4A). Notably, however, map3k1 RNAi animals did not regenerate all other ectopic eyes that were removed. This indicates that the TZ location is maintained after map3k1 RNAi, and that at least some progenitors are capable of migrating and differentiating at the normal TZ location in the absence of an eye. These data are consistent with a model in which map3k1 does not primarily control the positional information read by progenitors, but instead affects the ability of progenitors to differentiate at proper locations in response to a normal positional information system – a possibility explored further below.

map3k1 RNAi animals display tissue-specific regeneration at target zones.

A. Top, schematic depicts experimental design. Bottom panels show live image examples of control and map3k1 RNAi animal eye resections at d0, d1, and d10 post eye resection. Scale bar, 200μm. B. Diagram, pharynx resection. Control and map3k1 RNAi animals at d1 and d10 post resection; accompanied by FISH of d10 pharynx regeneration using probes to vitrin (pharynx-specific), NB.22.1e (mouth and esophagus), and mhc1 (pharynx muscle). Scale bar, 200μm; See also Figure 4- figure supplement 1. C. d10 regeneration in head, trunk, and tail fragments. An anterior-shifted pharynx regenerated in a map3k1 RNAi tail fragment. Scale bar, 200μm. D. FISH images of d10 trunk and tail regenerates showing regeneration of photoreceptors (anti-Arrestin), pharynx (vitrin), and mouth and esophagus (NB.22.1e). An anterior shift in vitrin+ cells in d10 tail regenerate. Scale bar, 200μm. E. AP distributions of original trunk pharynges in control and map3k1 RNAi animals and ectopic differentiation events in map3k1 RNAi trunk regenerates **** p <.0001 Mann-Whitney-U test. F. AP distributions of regenerated pharynges in tail fragments shows an anterior shift in pharynx regeneration in map3k1 RNAi animals compared to control animals. *** p <.0001 Mann-Whitney-U test.

map3k1 RNAi animals can undergo tissue-specific and whole-body regeneration, with some errors in pharynx organization.

A. FISH following pharynx resection at day zero (d0) with the markers vitrin (pharynx-specific), NB.22.1e (mouth and esophagus), and mhc1 (pharynx muscle). Scale bars, 200μm. B. Example of map3k1 RNAi d10 pharynx resection. FISH shows dorsal (left) and ventral (right) views of the same animal, showing disorganized pharynx regeneration in map3k1 RNAi animals. C. FISH of d10 head regeneration showing disorganized regeneration of the pharynx using vitrin and NB.22.1e RNA probes. Scale bars, 200μm. n=10/10. D. FISH showing dispersed notum+ cells in a d7 tail regenerate compared to control animals. Scale bar, 200μm

To determine if the pharynx target zone is also maintained in its normal, wild-type location in map3k1 RNAi animals, we removed the entire pharynx from its normal location during progression of the map3k1 RNAi phenotype (Figure 4B, Figure 4- figure supplement 1A). By 10 days after pharynx resection, animals regenerated a pharynx in the original pharynx position, as well as displayed an additional anterior pharynx-like structure (Figure 4B, Figure 4- figure supplement 1B). Thus, the case for the pharynx is more complex than for the eye. Regardless, these findings for the pharynx suggest that at least the normal central TZ remains and is not simply shifted anteriorly.

The location of de novo organ regeneration after map3k1 RNAi

To determine the location of progenitor targeting and de novo organ formation during regeneration, we amputated map3k1 RNAi animals into head, trunk, and tail fragments and analyzed them after 10 days of regeneration. All fragments regenerated any organs that they did not already contain at the time of amputation (Figure 4C, D). Two eyes formed in approximately normal locations in map3k1 RNAi head blastemas, rather than being posteriorly shifted or appearing in multiple locations initially. This suggests, similarly to the results from the eye resection experiments described above, that the TZ is regenerated in roughly the wild-type location in map3k1 RNAi animals.

Prior work in the planarian species D. japonica showed that map3k1 RNAi results in tail fragments regenerating pharynges in an anterior-shifted location (Hosoda et al., 2018). Consistent with this previously reported effect, pharynges regenerated more anteriorly in map3k1 RNAi S. mediterranea tail fragments, just posterior to the regenerating brain (Figure 4C-E). Notably, trunks also regenerated secondary pharynx-like aggregates very close to the brain that were underdeveloped, and appeared to interfere with structures around them (Figure 4D, E). Head fragments regenerated multiple pharyngeal structures (Figure 4- figure supplement 1C). Therefore, the situation for the pharynx in regeneration is more complex than that of the eye, similar to the findings for organ resections described above.

PCGs and pole organization are largely unaffected by map3k1 RNAi

As noted above, numerous PCGs can be inhibited to cause organ duplications. However, the map3k1 RNAi phenotype described so far is largely consistent with alteration of progenitor targeting behavior rather than global positional information shifting, as exemplified by the case of the eye. To directly examine the spatial maintenance of positional information in map3k1 RNAi animals we labeled these animals with RNA probes for multiple PCGs. Anterior, posterior, and medial PCG expression domains were largely unaffected in map3k1 RNAi animals – including expression domains for sFRP-1, ndl-5, ndl-2, ndl-3, wntP-2, axinB, and slit (Figure 5A, B). The notum+ anterior pole, which is produced from migratory FoxD+ progenitors (Scimone et al., 2014b), was present, but displayed some dispersal of cells, somewhat reminiscent of the eye phenotype (Figure 5B). The anterior pole also displayed some disorganization during head regeneration (Figure 4- figure supplement 1D). The posterior pole appeared normal (Figure 5B). axinB transcription can serves as a read-out of the posterior-to-anterior Wnt activity gradient in planarians (Iglesias et al., 2008; Reuter et al., 2015; Stückemann et al., 2017; Tewari et al., 2019). axinB was transcribed in a similar pattern in negative control and in map3k1 RNAi animals, indicating that Wnt activity was maintained in a relatively normal spatial distribution (Figure 5A). These findings are consistent with results described so far: if enough progenitors can still make it to the correct TZ location, eyes will be maintained and regenerated at the normal TZ, indicating positional information still remains and defines a TZ for the eye at approximately the correct location in map3k1 RNAi animals. The fact that ectopic eyes emerge over time in a disordered fashion suggests that it is progenitor targeting for differentiation at precise locations that is affected by map3k1 RNAi rather than positional information.

Positional information remains largely unaffected in map3k1 RNAi animals

A. FISH panel of position control genes (sFRP-1, ndl2, ndl5, ndl3, wntP-2, axinB, and slit) shows no obvious changes to positional information. Sample numbers are indicated in the lower left of each panel. Scale bar, 200μm. B. FISH showing some dispersion of notum+ cells in the heads of map3k1 RNAi animals compared to control animals. Right panels, no obvious changes in posterior pole organization. Sample numbers are indicated in the lower left of each panel, scale bars, 200μm.

map3k1 is expressed in neoblasts and post-mitotic progenitors.

A. map3k1 is expressed broadly across neoblast clusters. Plot points are ordered. B. map3k1 is expressed broadly across post-mitotic progenitors; plot points are ordered. C. map3k1 is expressed in some six-1/2+ neoblasts, which contain eye neoblasts; plot points are ordered. D. map3k1 is expressed in some G0 post-mitotic eye progenitors; plot points are ordered. E. map3k1 is expressed within the predicted progenitor cluster (Neural 1) of the affected cell types: dd_17258 and dd_16476; plot points are ordered. F. dd_17258 mature marker expression within its predicted progenitor cluster (Neural 1); plot points are ordered. G. dd_16476 mature marker expression within its predicted progenitor cluster (Neural 1); plot points are ordered. H. map3k1 expression in neural, parenchymal, pharyngeal, and six1/2+ neoblast clusters. I. map3k1 expression in neural, parenchymal, and pharyngeal post-mitotic progenitor clusters.

map3k1 is expressed in neoblasts and migratory post-mitotic progenitors

Recent scRNA-seq work has developed annotated fate-associated clusters of planarian neoblasts and post-mitotic progenitors (King et al., 2024). If map3k1 acts in progenitors to regulate their differentiation, is should be transcribed in these cells. Indeed, map3k1 transcripts were present broadly across neoblast and post-mitotic progenitor clusters (Figure 5- figure supplement 1), including for the eye (Figure 5- figure supplement 1C, D), neural classes affected by map3k1 RNAi (Figure 5- figure supplement 1E-G), and for the pharynx (Figure 5- figure supplement 1H, I). Whereas expression data alone do not necessarily map3k1, these data are consistent with the possibility that map3k1 can act within migratory progenitors.

map3k1 is required for restricting differentiation of eye progenitors along their migratory path to the target zone

Eye and pharynx progenitors are normally specified in restricted domains (Lapan and Reddien, 2011, 2012; Adler et al., 2014; Scimone et al., 2014a; Atabay et al., 2018), (referred to here as progenitor specification zones). Within these domains, progenitors migrate to reach their target tissue at their TZ, where they differentiate (Atabay et al., 2018; Hill and Petersen, 2018). We sought to test the possibility that the ectopic differentiated cells in map3k1 RNAi animals resulted from premature progenitor differentiation at positions on their normal migratory paths, before reaching the TZ. An alternative scenario we considered is that ectopic progenitor specification at some distant location occurs after map3k1 RNAi and requires excessively long-range progenitor migration to reach the TZ, ultimately resulting in ectopic differentiation.

Planarian neoblasts can be killed with irradiation (Bardeen and Baetjer, 1904; Reddien et al., 2005). Lead shielding can be utilized in X-irradiation experiments to locally protect neoblasts, resulting in neoblasts present only in a restricted field (Dubois, 1948; Guedelhoefer and Sánchez Alvarado, 2012; Abnave et al., 2017). Neoblasts expand only very slowly from these restricted regions (barring amputation) (Saló and Baguñà, 1985), but their postmitotic descendant cells serving as progenitors can readily migrate out from this region to differentiate at target tissues (Guedelhoefer and Sánchez Alvarado, 2012; Abnave et al., 2017; Park et al., 2023). We utilized lead shielding over the head and pre-pharyngeal region and X-irradiation to preserve neoblasts only in the anterior region where eye progenitors are primarily specified (Figure 6A). 48 hours after irradiation, we initiated RNAi of map3k1 to observe the behavior of any eye progenitors restricted to be born within their primary eye progenitor specification zone (the zone of surviving neoblasts). 17 days after the initiation of map3k1 RNAi, ectopic eye cells and ectopic dd_17258+ neurons were apparent (Figure 6B, C). This indicates that ectopic differentiation can occur after map3k1 RNAi even from neoblasts restricted to undergo fate specification close to their target tissue.

map3k1 RNAi eye progenitors prematurely differentiate along their typical migratory path.

A. Schematic of head-shielded irradiation experimental design. B. FISH showing instances of ectopic photoreceptor (anti-Arrestin), optic cup (catalase, tyrosinase, glut3), and dd_17258 cells born in the upper half of the animal (n=3/3 with good shielding and cells present correspond to area of leftover neoblasts). C. Schematic of EdU-labeled transplant experimental design. D. FISH and antibody staining showing examples of EdU+ eye cells from map3k1 RNAi animals ectopically differentiated in wild-type animals (n=7/18 with EdU+ ectopic cells; n=16/18 exhibiting any ectopic eye cells outside the transplant area). Scale bar, 50μm.

map3k1 RNAi animals display to improper progenitor targeting and teratoma formation.

A. FISH showing EdU+ photoreceptors (anti-Arrestin) in the eye of the recipient animal. These cells differentiated from map3k1 RNAi progenitors born during the EdU pulse of the donor animal, following wild-type migratory cues out of the graft and into the recipient eye. Scale bar, 200μm. B. FISH image of EdU-ectopic eye cells in recipient wild-type animals after EdU plug transplantation. Scale bar, 200μm. C. Live images of early growths that will eventually become teratomas in map3k1 RNAi animals. Scale bar, 100μm.

Second, we EdU-labeled neoblasts in map3K1 RNAi animals after 24 days of RNAi, then transplanted a pre-pharyngeal tissue fragment from these animals into the prepharyngeal region of unlabeled control host animals (hosts lack map3K1 RNAi) (Figure 6D). Animals were then fixed 12 days later. This experimental design allowed assessment of the behavior of map3k1 RNAi progenitors that migrate into a non-RNAi host environment, from within the normal eye progenitor specification zone. EdU-labeled cells migrated out of the transplantation region and differentiated into eye cells at both the normal eye location (at the TZ) and at ectopic locations in the host environment (Figure 6E). Additional ectopic eye cells were observed following transplantation that were not EdU-labeled, likely the result of additional mistargeting of map3k1 RNAi progenitors from the transplant, but lacking EdU because of incomplete EdU labeling or from cells born after the EdU pulse (Supplemental Figure 5A, B). These observations indicate that eye progenitors originating from the normal specification zone can erroneously differentiate before reaching their target location.

Differentiated cells in the wrong organ of map3k1 RNAi animals

We administered map3k1 dsRNA for RNAi and let animals undergo normal long-term tissue turnover to observe the consequences of errors in progenitor targeting. We observed differentiated pharynx muscle cells within the cephalic ganglia (Figure 7A). We also observed other differentiated pharynx and epidermal cell types within the eye (Figure 7B), including in regenerated animals (Supplemental figure 5C). These striking cell-organization defects were not previously observed in the patterning phenotypes observed following RNAi of PCGs. We suggest that this defect highlights the risk to tissue architecture of not enacting tight spatial regulation of differentiation, especially in a biological context where progenitors are spatially dispersed and migratory. Ectopic differentiated cells can become incorporated into inappropriate tissue environments where they would normally never be observed, potentially disrupting tissue structure and function.

map3k1 RNAi results in differentiation in incorrect organs and teratoma formation.

A. FISH of long term (>6 weeks) map3k1 RNAi animal showing clusters of mhc1 cells that appear to have formed small structures within the lobes of the brain and ventral cords. Right panels depict NB.22.1e cells incorporated into the eye (anti-Arrestin). Number of animals analyzed is indicated in the lower left of each panel; Left panel scale bar, 200μm; all other panels scale bars, 50μm. B. FISH showing vitrin+ cells incorporated into the eyes of a d10 tail regenerate. n=9/24 animals contained at least one vitrin+ cell in one eye; left panels scale bar, 200μm. All other scale bars, 50μm. C. Left panel contains live images of early developing growths in map3k1 RNAi animals, accompanied by a DAPI image of an outgrowth in between the brain lobes. Scale bars 200μm; The right panel contains live images and FISH examples of late-stage teratomas from map3k1 RNAi animals between 12 and 16 feedings, accompanied by FISH examples showing eyes dispersed around the teratomas. Scale bars, 200μm. D. Left panels are FISH images showing an exclusion of vitrin+ cells from lateral outgrowths but a common presence of NB.22.1e+ cells in outgrowths. mag1, a parenchymal marker with transcripts abundant in cells in the neck region is also mostly excluded from outgrowths. Right panel shows examples of cell types commonly found in outgrowths: cintillo (neuron), dd_17258 (neuron), dd_3534 (neuron), estrella (glia), anti-Arrestin (photoreceptor projections), NB.22.1e, lamin (mouth, epidermis), and col2 (muscle). Scale bar 100μm.

map3k1 RNAi animals develop teratomas

Long-term map3k1 RNAi in animals undergoing tissue turnover without injury led to the inevitable emergence of tissue growths (Figure 7C, D, Figure 6- figure supplement 1C). These growths predominantly formed in the anterior of the animal and presented as a heterogeneous collection of cell and tissue types. vitrin+ cells and gland cells (mag1) were not as common in lateral outgrowths. Cell types that were present in these teratomas included eye cells, various neuron types from the central nervous system (cintillo+, dd_17258+, dd_3524+), glia, muscle cells, and NB.22.1e cells (Figure 7D). A similar patterning defect has been observed in planarians with a defect in progenitor migration caused by integrin RNAi (Bonar and Petersen, 2017; Seebeck et al., 2017). It is known that differentiated tissues, such as the eye, can trap their own progenitors and lead to their differentiation (Atabay et al., 2018; Hill and Petersen, 2018). This suggests that ectopic differentiation in inappropriate locations in map3k1 RNAi animals can trap more progenitors and lead to an inappropriate aggregate of differentiated cells with aberrant pattern and organization. This highlights a further risk to the formation of tissue architecture if the differentiation of migratory progenitors is not tightly controlled spatially.

Discussion

Planarians display continuous turnover of adult tissues through the fate specification and differentiation of adult stem cells called neoblasts (Reddien, 2018). Fate specification in neoblasts can occur regionally (such as in the head for eyes), but is still spatially broad and intermingled. Fate-specified neoblasts (specialized neoblasts) produce progeny cells (post-mitotic progenitors) that migrate to precise locations for local, highly patterned differentiation. Progenitor targeting for local differentiation requires regulation from stem-cell-extrinsic signals, in the form of regionally expressed genes in muscle that comprise adult planarian positional information. How positional information interfaces with neoblasts and post-mitotic progenitors at the molecular level to regulate migratory assortment and differentiation only at precise locations is unknown and a fundamental problem of planarian regeneration and progenitor differentiation regulation. We sought factors that might be required for this process through RNAi-based gene perturbation experiments, and uncovered a requisite role in the spatial regulation of stem cell differentiation for the map3k1 gene, which encodes an ortholog of vertebrate MAP3K1. map3k1 RNAi animals displayed an abnormal pattern of progenitor differentiation, with differentiation occurring in disorganized locations along post-mitotic progenitor migratory paths. We suggest a model in which map3k1 is required for the spatial restriction of differentiation in stem cell lineages until suitable differentiation cues are encountered, either in the form of positional information (a target zone) or cues from target tissues (Figure 8A).

Model: Map3K1 restricts migratory progenitor differentiation until they reach their target.

A. Schematic showing the inhibition of differentiation of a specific progenitor until reaching its target tissue at the target zone. B. Defects observed with map3k1 RNAi, highlighting significance of restricting differentiation until suitable contextual cues are reached.

The defect in spatial restriction of progenitor differentiation following map3k1 RNAi can lead to dramatic tissue patterning defects, including the differentiation of ectopic isolated cells that normally only appear in groups (e.g., isolated photoreceptor neurons), the emergence of ectopic organs, differentiated cells from one tissue type being present within an incorrect organ, and teratoma formation (Figure 8B). These tissue mispatterning attributes highlight the significance of spatial regulation of differentiation during targeting of migratory progenitors for maintaining and regenerating adult pattern. This process might prove important development in many organisms, and be particularly important in the context of adult regeneration, where tissue scale can be large and adult progenitors at least in some contexts can be challenged to migrate large distances before differentiating.

There is some differentiation of eye progenitors in the tail of map3k1 RNAi planarians, far from the normal location of eye progenitor production and differentiation. However, only a very low rate of eye progenitor production was observed in the tail of map3k1 RNAi animals, and a comparable low frequency of eye-specialized neoblasts were present in control tails. One interesting possibility that can be investigated in the future is that there is some noise in the spatial regulation of stem cell fate specification in planarians that does not normally contribute to errors in the spatial distribution of differentiated cells because of prevention of differentiation until cells reach their appropriate location. The progenitor specification domain for pharynx cells, by contrast, appeared to expand anteriorly after map3k1 RNAi, indicating a possible role in the spatial regulation of progenitor specification in some cases.

Several observations suggest that the defects in targeting in map3k1 RNAi animals involves a process within the neoblasts and their post-mitotic progenitor descendants, rather than in a role for map3k1 in establishing the progenitor-extrinsic cues (e.g., muscle PCGs) that facilitate progenitor targeting. First, PCG expression patterns themselves were largely normal following map3k1 RNAi. Second, ectopic differentiation after map3k1 RNAi was more spatially disorganized than is observed for patterning phenotypes that occur following the shifting of PCG coordinates (such as a posterior line of ectopic eyes in ndk RNAi animals (Cebria et al., 2002) or a lateral line of ectopic eyes in wnt5 RNAi animals (Atabay et al., 2018)). Third, ectopic cells following map3k1 RNAi were frequently isolated, as opposed to appearing exclusively in organized aggregates or forming ectopically placed organs. Fourth, transplantation of EdU-labeled tissue grafts from map3k1 RNAi animals into the pre-pharyngeal region of control animals showed instances of ectopic differentiation in wild-type host tissue. Finally, during regeneration or following eye resection, progenitors can still be targeted to the correct target zone for the eye, indicating that the eye target zone still remains after map3k1 RNAi. One possibility is that Map3k1 applies a brake on differentiation in progenitors that is released or overcome by some cue associated with a progenitor reaching its target zone or by interacting with its target tissue.

Map kinase signaling cascades are versatile pathway modules found throughout the animal kingdom and are involved in diverse processes (Widmann et al., 1999). The kinases and accompanying cofactors in these modules enable the regulated activation of various Map kinases (e.g., ERK1/2, p38, JNK) (Suddason and Gallagher, 2015). Map kinase kinase kinases (MAP3Ks) are one of the first activated proteins in these cascades, often responding to receptor tyrosine kinase signaling at the cell membrane. In mammals, there are 24 characterized MAP3Ks: MAP3K1 through MAP3K21, B-Raf, C-Raf, and A Raf, which activate downstream MAP2K proteins through phosphorylation. Among the MAP3K proteins, MAP3K1 is unique in its domain composition. In particular, MAP3K1 orthologs are the only MAP3K with a PHD domain (Suddason and Gallagher, 2015). The MAP3K1 domain architecture enables a role in ubiquitination as well as kinase activity (Pham et al., 2013). In mammals, map3k1 has a role in regulating many processes such as proliferation, differentiation, division, and cell death (Suddason and Gallagher, 2015). map3k1 gene functions are not well characterized in most invertebrate systems, and Drosophila and C. elegans have no identified map3k1 orthologs (Widmann et al., 1999). Planarians therefore present an attractive invertebrate model for dissection of map3k1 function. There is evidence for the role of various Map-kinases (e.g., ERK, MEK, RAS, p38) in planarian regeneration, particularly in the blastema formation and the wound and missing tissue response programs (Tasaki et al., 2011a; Tasaki et al., 2011b; Owlarn et al., 2017; Wang et al., 2020). map3k1 has been studied in a different planarian species, D. japonica (Hosoda et al., 2018). This work suggested a possible role for map3k1 in the scaling and patterning of the trunk and head regions of regenerating animals, proposed to be through alternation of ERK and β-catenin gradients (Hosoda et al., 2018). Additionally, map3k1 has been implicated in germ cell proliferation and terminal differentiation of stem cells in the parasitic flatworm E. multilocularis through JNK signaling (Stoll et al., 2021). It will be of interest to further dissect the molecular role of map3k1 in planarian progenitor differentiation, and to determine whether map3k1 orthologs have similar roles in regulating differentiation in additional regenerative organisms.

Adult patterning systems in some organisms can rely on a spatially coarse progenitor specification system, migratory assortment of progenitors, and local differentiation. This process involves progenitor transitions from spatially broad and disorganized, to local and highly patterned as cells transit from stem cell to post-mitotic migratory progenitor. This suggests that regulation of terminal differentiation must have a central role in pattern production. Progenitor-extrinsic positional information has a prominent regulatory role in this process, and some mechanism(s) could exist within the progenitors themselves to enable spatial restriction of differentiation until suitable cues have been received. We suggest that map3k1 acts within planarian progenitors to mediate such spatial restriction on differentiation, and that this is critical for preventing mistargeting differentiation to the incorrect location or organ, and to prevent teratoma formation.

Materials and methods

Animal husbandry and surgery

Asexual S. mediterranea clonal strain CIW4 was used for all experiments. Animals were cultured in static 1x Montjuic water (1.6 mmol/l NaCl, 1.0 mmol/l CaCl2, 1.0 mmol/l MgSO4, 0.1 mmol/l MgCl2, 0.1 mmol/l KCl and 1.2 mmol/l NaHCO3 prepared in Milli-Q water) at 20°C. Amputations were performed under cold conditions (∼4C) with a scalpel. Animals were fed homogenized beef liver weekly, with water changed biweekly. Animals were starved for approximately seven days before experiments.

Whole-mount fluorescent in situ hybridization (FISH)

Animal mucus was removed using 5% N-acetylcysteine in PBS; animals were then fixed with 4% formaldehyde in PBST for 20 minutes, with rocking. Animals were then washed with PBST, incubated in 1:1 PBST:methanol, then stored in 100% methanol at −20C until ready for bleaching. Animals were moved into mesh baskets in a 24-well plate where all remaining steps were carried out. Animals were placed on a light source to bleach for 1.5 hours in a bleaching solution (5% formamide, 0.5x SSC, and 1.2% hydrogen peroxide). After 2 PBST washes, animals were then treated with 5 mg/ml Proteinase K for 10 minutes, followed by 4% formaldehyde post-fixation in PBST.

RNA probes were transcribed using DIG, FITC, or DNP-modified nucleotides, allowing signal amplification through the use of DIG, FITC, and DNP antibodies conjugated to an HRP. Probes were diluted in Hybe solution (1:800) (50% deionized formamide, 5x SSC, 1 mg/mL yeast RNA, 1% Tween-20, 5% dextran sulfate), and left to incubate overnight. The following days, we performed antibody incubations at 4C overnight using anti-DIG-POD (1:1500, Roche; 10% western blocking solution (Roche)), anti-FITC-POD (1:2,000, Roche; 5% horse serum, 5% western blocking solution), and anti-DNP-HRP (1:100, Perkin-Elmer; blocking solution with 10% inactivated Horse Serum). Tyramide signal amplification involved incubating in rhodamine (1:1,000), fluorescein (1:1,500), or Cy5 (1:300) in borate buffer (0.1M boric acid, 2M NaCl, pH 8.5) containing 0.0003% hydrogen peroxide for 10 minutes. Samples were incubated in 1% sodium azide for 2 hours to inactivate the HRP. Blocking and antibody incubations then occurred for detection of the second probe. Animals were incubated overnight in 1mg/mL DAPI solution at 4C. Animals were mounted on coverslips in ProLong Gold Antifade Mountant (Thermo Fisher).

EdU labeling and detection

F-ara-EdU (Click Chemistry Tools) was diluted in Dimethyl sulfoxide (DMSO) to 200mg/ml, then diluted in static 1x Montjuic water to 1.25mg/mL. Animals were split into 10 animals per well in a 12-well plate, then soaked in 1.25mg/mL EdU solution for 20 hours following 1 week of starvation. EdU solution was replaced with 5mg/mL Instant Ocean sea salt dissolved in Milli-Q water. Prior to probe hybridization in the in situ hybridization protocol, following proteinase K and 4% formaldehyde incubations, cells were incubated in a “click reaction”-1% 100 mM CuSO4, 0.1% 10 mM TAMRA-Azide-fluor 545 (Sigma-Aldrich), and 20% 50 mM ascorbic acid in PBS- for 30 minutes in the dark, proceeded by 6 PBST washes and continuation of the probe hybridization step.

RNA interference

dsRNA for RNAi was prepared using in vitro transcription with T7 polymerase (Promega). Template was generated with PCR; primers contained T7 promoters. RNA was resuspended in water and dsRNA annealed. C. elegans unc-22 dsRNA was used as the negative control for all RNAi experiments.

Transplantation

EdU plug transplants were performed using map3k1 RNAi animals, 12 hours following a 20 hour EdU pulse, as the donor to a recipient wild type animal. Donor animals were anesthetized with 0.2% chlorotone solution, followed by an incubation in Holfreter’s solution, then placed on an ice block covered in Whatman filter paper moistened with 1x Montjuic water to surgically manipulate with a clean scalpel. EdU+ map3k1 RNAi donor animals and wild type recipients both had a center portion of their pre-pharyngeal regions removed. The pre-pharyngeal donor graft from the map3k1 RNAi animal was placed in the EdU-wild type recipient’s pre-pharyngeal region. Recipients were then gently covered with cigarette paper soaked in chilled Holfreter’s solution and transferred to a small petri dish with just enough Holfreter’s solution to cover the bottom of the dish. Animals were put at 10C for 20 hours; the following day, transplant recipients were gently recovered and put into 1x Montjuic water containing .1% gentamicin (Gibco).

Shielded irradiation

Animals were irradiated using a Gammacell-40 137-cesium source. For shielded irradiation experiments, animals were anesthetized with 0.2% chlorotone, then arranged on Whatman filter paper in a petri dish sitting on ice. Animals were oriented to have their anterior half covered by the lead shield placed over the petri dish. Samples were placed in a Gammacell chamber and exposed to 3000 Rad of unidirectional X-irradiation. Animals were rescued with 1x Montjuic water and stored in 1x Montjuic water with .1% gentamicin (Gibco) until further experimentation.

Image analysis and statistical analysis

For cell counts, the AP axis was binned into 6 regions according to anatomical landmarks. AP_1 (head tip to bottom of brain), AP_2 (bottom of brain to top of pharynx), AP_3 (top of pharynx to middle of pharynx), AP_4 (middle of pharynx to bottom of pharynx), AP_5 (top half of the tail) and AP_6 (bottom half of the tail). Each animal had one data point for each of the 6 bins.

Graphpad Prism software was used for graph generation and statistical analysis using a Mann-Whitney-U test with a significance cut off of p=.05. For cell count comparisons in the tail, Graphpad Prism software was used for graph generation and statistical analysis using a student t-test with a significance cut off of p=.05.

eLife Assessment

This valuable study examines the role of map3k1, a MAP3K family member that has both kinase and ubiquitin ligase domains, in the differentiation of progenitors in the flatworm Planaria. The convincing analyses demonstrate that map3k1 acts within progenitors to restrict their premature differentiation and to prevent formation of teratomas. This work would be of interest to researchers in the fields of regeneration, developmental biology, and aging.

Reviewer #1 (Public review):

Summary:

The authors assess the role of map3k1 in adult Planaria through whole body RNAi for various periods of time. The authors' prior work has shown that neoblasts (stem cells that can regenerate the entire body) for various tissues are intermingled in the body. Neoblasts divide to produce progenitors that migrate within a "target zone" to the "differentiated target tissues" where they differentiate into a specific cell type. Here the authors show that map3k1-i animals have ectopic eyes that form along the "normal" migration path of eye progenitors (Fig. 1), ectopic neurons and glands along the AP axis (Fig. 2) and pharynx in ectopic anterior positions (Fig. 3). The rest of the study show that positional information is largely unaffected by loss of map3k1 (Fig. 4,5). However, loss of map3k1 leads to premature differentiated of progenitors along their normal migratory route (Fig. 6). They also show that an ill-defined "long-term" whole body depletion of map3k1 results in mis-specified organs and teratomas.

Strengths:

(1) The study has appropriate controls, sample sizes and statistics.

(2) The work appears to be high-quality.

(3) The conclusions are supported by the data.

(4) Planaria is a good system to analyze the function of map3k1, which exists in mammals but not in other invertebrates.

Weaknesses:

(1) The paper is largely descriptive with no mechanistic insights.

(2) Given the severe phenotypes of long-term depletion of map3k1, it is important that this exact timepoint is provided in the methods, figures, figure legends and results.

(3) Fig. 1C, the ectopic eyes are difficult to see, please add arrows.

(4) line 217 - why does the n=2/12 animals not match the values in Fig. 3B, which is 11/12 and 12/12. The numbers don't add up. Please correct/explain.

(5) Figure panels do not match what is written in the results section. There is no Fig. 6E. Please correct.

Reviewer #2 (Public review):

Summary:

The flatworm planarian Schmidtea mediterranea is an excellent model for understanding cell fate specification during tissue regeneration and adult tissue maintenance. Planarian stem cells, known as neoblasts, are continuously deployed to support cellular turnover and repair tissues damaged or lost due to injury. This reparative process requires great precision to recognize the location, timing, and cellular fate of a defined number of neoblast progeny. Understanding the molecular mechanisms driving this process could have important implications for regenerative medicine and enhance our understanding of how form and function are maintained in long-lived organisms such as humans. Unfortunately, the molecular basis guiding cell fate and differentiation remains poorly understood.

In this manuscript, Canales et al. identified the role of the map3k1 gene in mediating the differentiation of progenitor cells at the proper target tissue. The map3k1 function in planarians appears evolutionarily conserved as it has been implicated in regulating cell proliferation, differentiation, and cell death in mammals. The results show that the downregulation of map3k1 with RNAi leads to spatial patterning defects in different tissue types, including the eye, pharynx, and the nervous system. Intriguingly, long-term map3k1-RNAi resulted in ectopic outgrowths consistent with teratomas in planarians. The findings suggest that map3k1 mediates signaling, regulating the timing and location of cellular progenitors to maintain correct patterning during adult tissue maintenance.

Strengths:

The authors provide an entry point to understanding molecular mechanisms regulating progenitor cell differentiation and patterning during adult tissue maintenance.

The diverse set of approaches and methods applied to characterize map3k1 function strengthens the case for conserved evolutionary mechanisms in a selected number of tissue types. The creativity using transplantation experiments is commendable, and the findings with the teratoma phenotype are intriguing and worth characterizing.

Weaknesses:

The article presents a provocative idea related to the importance of positional control for organs and cells, which is at least in part regulated by map3k1. Nonetheless, the role of map3k1 or its potential interaction with regulators of the anterior-posterior, mediolateral axes, and PCGs is somewhat superficial. The authors could elaborate or even speculate more in the discussion section and the different scenarios incorporating these axial modulators into the map3k1 model presented in Figure 8.

The article can be improved by addressing inconsistencies and adding details to the results, including the main figures and supplements. This represents one of the most significant weaknesses of this otherwise intriguing manuscript. Below are some examples of a few figures, but the authors are expected to pay close attention to the remaining figures in the paper.

Details associated with the number of animals per experiment, statistical methods used, and detailed descriptions of figures appear inconsistent or lacking in almost all figures. In some instances, the percentage of animals affected by the phenotype is shown without detailing the number of animals in the experiment or the number of repeats. Figures and their legends throughout the paper lack details on what is represented and sometimes are mislabeled or unrelated. Specifically, the arrows in Figure 1A are different colors. Still, no reasoning is given for this, and in the exact figure, the top side (1A) shows the percentages and the number of animals below. Conversely, in Figures 1B, C, and D, no details on the number of animals or percentages are shown, nor an explanation of why opsin was used in Figure 1A but not 1B. Is Figure 1B missing an image for the respective control? Figure 1C needs details regarding what the two smaller boxes underneath are. Figure 1C could use an AP labeling map in 10 days (e.g., AP6 has one optic cup present). Figure 1C and F counts do not match. In Figure 1C, we do not know the number of animals tested, controls used, the scale bar sizes in the first two images, nor the degree of magnification used despite the pharynx region appearing magnified in the second image. Figure 1C is also shown out of chronological order; 36 days post RNAi is shown before 10 days post RNAi. Moreover, the legends for Figures 1C and 1D are swapped.

Additionally, Figure 1F and many other figures throughout the paper lack overall statistical considerations. Furthermore, Figure 1F has three components, but only one is labeled. Labeling each of them individually and describing them in the corresponding figure legend may be more appropriate.

Figure 2C shows images of gene expression for two genes, but the counts are shown for only one in Figure 2D. It is challenging to follow the author's conclusions without apparent reasoning and by only displaying quantitative considerations for one case but not the other. These inconsistencies are also observed in different figures. In Figure 2D, 24/24 animals were reported to show the phenotype, but only eight were counted (is there a reason for this?). In Figure 2E, the expression for three genes is shown, with some displaying anterior and posterior regions while others only show the anterior picture. Is there a particular reason for this? Also, in Figure 2F, the counts are shown for only the posterior region of two genes out of the three displayed in Figure 2E. It is unclear why the authors do not show counts for the anterior areas considered in Figure 2E. Furthermore, the legend for Figure 2D is missing, and the legend for 2F is mislabeled as a description for Figure 2D.

Supplement Figure 1 B reports data up to 6 weeks, but no text in the manuscript or supplement mentions any experiment going up to 6 weeks. There are no statistics for data in Supplement Figure 1E. Any significance between groups is unclear.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors assess the role of map3k1 in adult Planaria through whole body RNAi for various periods of time. The authors' prior work has shown that neoblasts (stem cells that can regenerate the entire body) for various tissues are intermingled in the body. Neoblasts divide to produce progenitors that migrate within a "target zone" to the "differentiated target tissues" where they differentiate into a specific cell type. Here the authors show that map3k1-i animals have ectopic eyes that form along the "normal" migration path of eye progenitors (Fig. 1), ectopic neurons and glands along the AP axis (Fig. 2) and pharynx in ectopic anterior positions (Fig. 3). The rest of the study show that positional information is largely unaffected by loss of map3k1 (Fig. 4,5). However, loss of map3k1 leads to premature differentiated of progenitors along their normal migratory route (Fig. 6). They also show that an ill-defined "long-term" whole body depletion of map3k1 results in mis-specified organs and teratomas.

Strengths:

(1) The study has appropriate controls, sample sizes and statistics.

(2) The work appears to be high-quality.

(3) The conclusions are supported by the data.

(4) Planaria is a good system to analyze the function of map3k1, which exists in mammals but not in other invertebrates.

Weaknesses:

(1) The paper is largely descriptive with no mechanistic insights.

The mechanistic insights we aim to address are primarily at the cellular systems level – how adult progenitor cells produce pattern. Specifically, we uncovered strong evidence that regulation of differentiation is an active process occurring in migratory progenitors and that this regulation is a major component of pattern formation during the adult processes of tissue turnover and regeneration. The map3k1 phenotype provided a tool used to reveal the existence of this regulation, and to understand the patterning abnormalities prevented by this regulatory mechanism. We updated the text in several places to make clearer some of this emphasis. For example, in the Discussion: "We suggest that differentiation is restricted during migratory targeting as an essential component of pattern formation, with the map3k1 RNAi phenotype indicating the existence and purpose of this element of patterning."

Naturally, identifying a particular molecule involved in this process is of interest for understanding molecular mechanism; this would allow for comparison to other cellular systems in other organisms and would focus future molecular inquiry. Future molecular studies into the mechanism of Map3k1 regulation and its downstream signaling will be fascinating as next steps towards understanding the process at the molecular level more deeply. We also added some discussion considering the types of upstream activation cues that could potentially be associated with Map3k1 regulation to suppress differentiation.

(2) Given the severe phenotypes of long-term depletion of map3k1, it is important that this exact timepoint is provided in the methods, figures, figure legends and results.

We removed the use of the term “long-term” and instead added timepoints used to all figure legends. We also added a summary of timepoints used in the methods section and included RNAi timepoint labels in figures where a phenotype progression over time is relevant to interpretation. For timecourses, we also added suitable time information to text in the results.

(3) Figure 1C, the ectopic eyes are difficult to see, please add arrows.

To improve visualization, we replaced the example animal in the original Figure 1C with one that has a stronger phenotype, including arrows pointing to every ectopic event. Additionally, we included magnified images of optic cup cells and photoreceptor neurons in the trunk and tail region. This is now Figure 1B.

(4) line 217 - why does the n=2/12 animals not match the values in Figure 3B, which is 11/12 and 12/12. The numbers don't add up. Please correct/explain.

In Figure 3B in the submitted version (3/18 had cells in the tail) had more animals scored (6 animals from a replicate experiment where 1/6 showed the cells in the tail) than the total scored (2/12 had cells in the tail) in the text, which did not have the animals from the replicate added during writing. The results are the same, just different sample sizes were noted in those locations and we fixed this issue. In the updated Figure 3, the order of presentation has shifted (e.g., prior 3B is now in 3C and Figure 3_figure supplement 1). We made sure to include numbers to all figure panels.

(5) Figure panels do not match what is written in the results section. There is no Figure 6E. Please correct.

Thank you for catching this. We have gone through figures and text after editing to make sure that text callouts are appropriately matched to the figures.

Reviewer #2 (Public review):

Summary:

The flatworm planarian Schmidtea mediterranea is an excellent model for understanding cell fate specification during tissue regeneration and adult tissue maintenance. Planarian stem cells, known as neoblasts, are continuously deployed to support cellular turnover and repair tissues damaged or lost due to injury. This reparative process requires great precision to recognize the location, timing, and cellular fate of a defined number of neoblast progeny. Understanding the molecular mechanisms driving this process could have important implications for regenerative medicine and enhance our understanding of how form and function are maintained in long-lived organisms such as humans. Unfortunately, the molecular basis guiding cell fate and differentiation remains poorly understood.

In this manuscript, Canales et al. identified the role of the map3k1 gene in mediating the differentiation of progenitor cells at the proper target tissue. The map3k1 function in planarians appears evolutionarily conserved as it has been implicated in regulating cell proliferation, differentiation, and cell death in mammals. The results show that the downregulation of map3k1 with RNAi leads to spatial patterning defects in different tissue types, including the eye, pharynx, and the nervous system. Intriguingly, long-term map3k1-RNAi resulted in ectopic outgrowths consistent with teratomas in planarians. The findings suggest that map3k1 mediates signaling, regulating the timing and location of cellular progenitors to maintain correct patterning during adult tissue maintenance.

Strengths:

The authors provide an entry point to understanding molecular mechanisms regulating progenitor cell differentiation and patterning during adult tissue maintenance.

The diverse set of approaches and methods applied to characterize map3k1 function strengthens the case for conserved evolutionary mechanisms in a selected number of tissue types. The creativity using transplantation experiments is commendable, and the findings with the teratoma phenotype are intriguing and worth characterizing.

Thank you to the reviewer for the positive feedback

Weaknesses:

The article presents a provocative idea related to the importance of positional control for organs and cells, which is at least in part regulated by map3k1. Nonetheless, the role of map3k1 or its potential interaction with regulators of the anterior-posterior, mediolateral axes, and PCGs is somewhat superficial. The authors could elaborate or even speculate more in the discussion section and the different scenarios incorporating these axial modulators into the map3k1 model presented in Figure 8

First, to strengthen the support for our finding that positional information is largely unaffected in map3k1 RNAi animals, we added data regarding the expression of additional relevant position control genes (PCGs) –ndl-4, ptk7, sp5, and wnt11-1 – to the PCG panel in Figure 5. The expression domain of ndl-4, an FGF receptor-like protein family member that contributes to head patterning and anterior pole maintenance, was normal in map3k1 RNAi. wnt11-1, a PCG with expression concentrated in the posterior end of the animal and with expression dependent on general Wnt activity, was also normal in map3k1 RNAi animals. ptk7, RNAi of which can result in supernumerary pharynges, also showed normal expression in map3k1 RNAi animals. Finally, sp5, a Wnt-activated gene with expression in the tail, also showed normal expression in map3k1 RNAi animals.

Second, to further support the conclusion that cells are not suitably responding to positional information after map3k1 RNAi, which we argue normally dictates where differentiation should occur, we added examples of differentiated cell types that are ectopically positioned within an atypical PCG expression domain for that cell type (Figure 5C). This underscores that following map3k1 RNAi the PCG expression domains do not change, but the pattern of differentiated cell types relative to these domains does shift. We also added data showing that regenerating tails had a proper wntP-2 gradient, but an anterior regenerating pharynx appeared outside of this wntP-2+ zone and inside of an ndl-5+ zone (Figure 5- figure supplement 1E). We added some discussion of these new data in the Figure 5 results section. We also noted, regarding independent recent map3k1 work (Lo, 2025), some evidence exists that a minor posterior shift in ndl-5 expression can occur after map3k1 RNAi.

Next, we added a new element to the model figure (Figure 8B) depicting that PCG expression domains remain normal after map3k1 RNAi, with ectopic differentiation occurring in an incorrect positional information environment. We refer to this new panel in the discussion: "We suggest that map3k1 is not required for the spatial distribution of progenitor-extrinsic differentiation-promoting cues themselves, but for progenitors to be restricted from differentiating until these cues are received (Figure 8B)."; we then follow this statement with a summary in the Discussion of six pieces of evidence that support this model.

Finally, we added some additional text to the discussion section about candidate mechanisms by which extrinsic cues could potentially regulate Map3k1, pointing to potential future inquiry directions. We suggest that map3k1 is not involved in regulating PCG activity domains themselves, but instead acts as a brake on differentiation within migratory progenitors through active signaling. This brake is then lifted when the progenitors hit their correct PCG-based migratory target, or when they hit their target tissue. How that occurs mechanistically is unknown. One scenario is that each progenitor is tuned to respond to a particular PCG-regulated environment (such as a particular ECM or signaling environment) to generate a molecular change that inactivates Map3K1 signaling, such as by inactivating or disengaging an RTK signal. Alternatively, the migratory process in progenitors could engage the Map3K1 signal, enabling signal cessation with arrival at a target location. When Map3K1 is active it could result in a transcriptional state that prevents full expression of differentiated factors required for maturation, tissue incorporation, and cessation of migration. These considerations are now added to the discussion.

The article can be improved by addressing inconsistencies and adding details to the results, including the main figures and supplements. This represents one of the most significant weaknesses of this otherwise intriguing manuscript. Below are some examples of a few figures, but the authors are expected to pay close attention to the remaining figures in the paper.

Details associated with the number of animals per experiment, statistical methods used, and detailed descriptions of figures appear inconsistent or lacking in almost all figures. In some instances, the percentage of animals affected by the phenotype is shown without detailing the number of animals in the experiment or the number of repeats. Figures and their legends throughout the paper lack details on what is represented and sometimes are mislabeled or unrelated.

We endeavored to ensure that these noted details are present throughout the legends and figures for all figure panels.

Specifically, the arrows in Figure 1A are different colors. Still, no reasoning is given for this, and in the exact figure, the top side (1A) shows the percentages and the number of animals below.

The only reason for the different colored arrows was for visibility purposes. To avoid confusion, we now use white arrows for all FISH images in figure 1, and where ever else possible. We also replaced the percentages with n numbers in the bottom left corner of the live images in Figure 1A.

Conversely, in Figures 1B, C, and D, no details on the number of animals or percentages are shown, nor an explanation of why opsin was used in Figure 1A but not 1B.

The original Figure 1B represented a few different examples of ectopic eye/eye cell patterns in the map3k1 RNAi animals to demonstrate the variable and disorganized nature of the phenotype. To address this, we added further explanation in the legend. We also merged 1A and 1B for simplicity of interpretation. opsin was used in Figure 1A to label cell bodies of photoreceptors. anti-Arrestin was used in the example FISH images to see if these cells were interconnected via projections, which we now clarify in the legend and in the text.

Is Figure 1B missing an image for the respective control? Figure 1C needs details regarding what the two smaller boxes underneath are.

The control for Figure 1B was in Figure 1A; the merger of Figures 1A/B should address this. Boxes in Figure 1C were labelled with numbers corresponding to the image above them.

Figure 1C could use an AP labeling map in 10 days (e.g., AP6 has one optic cup present). Figure 1C and F counts do not match.

We added a cartoon to 1C to show the region imaged. Note that the 36d trunk image (now Fig. 1B) has now been replaced with a full animal image and magnified boxes that show locations of example ectopic cells. That cell in 1C was categorized as in AP5. Note that additional animals were analyzed and data added to the distribution (now Fig. 1D).

In Figure 1C, we do not know the number of animals tested, controls used, the scale bar sizes in the first two images, nor the degree of magnification used despite the pharynx region appearing magnified in the second image. Figure 1C is also shown out of chronological order; 36 days post RNAi is shown before 10 days post RNAi. Moreover, the legends for Figures 1C and 1D are swapped.

We have endeavored to ensure sample numbers, control images, and appropriate scale bar notation in legends are present for all images. Figure 1C has now been split into two panels: Figure 1B and Figure 1C. It does not follow a chronological order in presentation for the following logic flow: we uncover and describe the phenotype; then, with knowledge of the defect, we walk back to see how early the phenotype starts after RNAi and what the pattern of ectopic cell distribution is when the phenotype starts to emerge (using the knowledge of which cells are affected from the overt phenotype described in 1A/B).

Additionally, Figure 1F and many other figures throughout the paper lack overall statistical considerations. Furthermore, Figure 1F has three components, but only one is labeled. Labeling each of them individually and describing them in the corresponding figure legend may be more appropriate.

The main point of the graphs in 1F (now 1D) was the overt overall pattern difference with the wild-type, which never has ectopic eye cells in the midbody or tail, and that the ectopic eye cells progress throughout the entire AP axis. However, we concur that a statistical test is a reasonable thing to show here and that is now included in the legend. The 3 components (in Figure 1F, now Figure 1D) where kept together with one figure label (D) for simplicity, but were rearranged (top and bottom) with a cartoon to the side and with modified labeling for extra clarity.

Figure 2C shows images of gene expression for two genes, but the counts are shown for only one in Figure 2D. It is challenging to follow the author's conclusions without apparent reasoning and by only displaying quantitative considerations for one case but not the other. These inconsistencies are also observed in different figures.

In Figure 2C, FISH images of cintillo+ and dd_17258+ neurons are shown to display the specificity of this effect to some neurons and not others. Because cintillo+ cells did not expand at all (n=24/24 animals), the counts for them would all be zero values. We only counted data for dd_17258 cells because it was the neuron that expanded compared to the control animals. We have now added a note in the legend explaining this.

In Figure 2D, 24/24 animals were reported to show the phenotype, but only eight were counted (is there a reason for this?).

8 animals were used to quantitatively characterize the spread of cells along the AP axis, as it was deemed an adequate sample size to capture the change in distribution of 17258+ cells from being head restricted to being present throughout the body. Through multiple cohorts of animals in replicates, a total of 24/24 examined animals showed this expansion phenotype. Double FISH experiments were additionally carried out using dd_17258 and various PCGs; these data are now included in Figure 5C, and these animals were added to the total counts regarding quantitative analysis of the phenotype in Figure 2D.

In Figure 2E, the expression for three genes is shown, with some displaying anterior and posterior regions while others only show the anterior picture. Is there a particular reason for this?

The original first panel in Figure 2E showed an example of a non-expanding gland cell type, dd_9223, which is very restricted to the head in both control and map3k1 RNAi animals. Because we did not observe a phenotype for this cell type (no cells in all control and map3k1 RNAi animal tails), we only included tail images of cell types that showed an abnormal phenotype with clear expanded to the posterior (dd_8476 and dd_7131). However, we have now included tail images of dd_9223 cells and added data for dd_9223 to the graph in Figure 2E.

Also, in Figure 2F, the counts are shown for only the posterior region of two genes out of the three displayed in Figure 2E. It is unclear why the authors do not show counts for the anterior areas considered in Figure 2E. Furthermore, the legend for Figure 2D is missing, and the legend for 2F is mislabeled as a description for Figure 2D.

We now include tail images for dd_9223 in Figure 2E to show that there are no ectopic cells in tails. We did not originally include counts of dd_9223 because there was no phenotype observed. dd_7131 and dd_8476 cell types appeared in the posterior of even control animals at a low frequency, unlike dd_9223 cells. However, we did now add counts for dd_9223 tail regions in the graph. We did not count the anterior regions of the animal because our goal was to show data for the visible phenotype (ectopic cells in the tail) not only with an example image, but also by showing the number of cells in the tail with a graph and statistical test. Legends have been updated with correct details.

Supplement Figure 1 B reports data up to 6 weeks, but no text in the manuscript or supplement mentions any experiment going up to 6 weeks. There are no statistics for data in Supplement Figure 1E. Any significance between groups is unclear.

More details about the RNAi feeding schedules have been added in the methods section. All RNAi timepoints are now specified specifically in the legends. The Figure 1F and Figure 1- figure supplement 1E (additional data: ovo+; smedwi-1- cell counts) and legends now mention the statistical tests performed and annotations (not significant *ns) or p values have been added to the graphs. For simplicity, we decided to include all smedwi-1+ counts together rather than splitting them into low and high smedwi-1+ cells, because we weren't really making any claims about low and high cells.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

It would be good to acknowledge in the discussion the recent paper from the Petersen lab on map3k1, published in PLoS Genet 2025, especially if the results differ between the two labs.

We added reference/discussion regarding the recent PLoS Genetics Lo, 2025 map3k1 paper at several suitable points in the manuscript.

Reviewer #2 (Recommendations for the authors):

Please pay close attention to the description of experimental details and the consistency throughout the paper. It seems like the reader has to assume or come across information that is not readily available from the text or the legends in the paper. This is an interesting paper with intriguing findings. However, the version presented here appears rushed or put together on the flight.

Thank you for your thorough feedback. We have endeavored to ensure all appropriate details are present in figures and/or figure legends.

Version 2

Abstract

Planarian regeneration and tissue turnover involve fate specification in pluripotent stem cells called neoblasts. Neoblasts select fates through the expression of fate-specific transcription factors (FSTFs), generating specialized neoblasts. Specialized neoblasts are spatially intermingled and can be dispersed broadly, frequently being present far from their target tissue. The post-mitotic progeny of neoblasts, serving as progenitors, can migrate and differentiate into mature cell types. Pattern formation is thus strongly influenced by the migratory assortment and differentiation of fate-specified progenitors in precise locations, which we refer to as progenitor targeting. This central step of pattern maintenance and formation, however, is poorly understood. Here, we describe a requirement for the conserved map3k1 gene in targeting, restricting post-mitotic progenitor differentiation to precise locations. RNAi of map3k1 causes ectopic differentiation of eye progenitors along their migratory path, resulting in dispersed, ectopic eye cells and eyes. Other neural tissues similarly display ectopic posterior differentiation and ectopic pharynx cells emerge dispersed laterally and anteriorly in map3k1 RNAi animals. Ectopic differentiated cells are also found within the incorrect organs after map3k1 RNAi, and ultimately, teratomas form. These findings implicate map3k1 signaling in controlling the positional regulation of progenitor behavior – restricting progenitor differentiation to targeted locations in response to external cues in the local tissue environment.

Introduction

Planarians are capable of regenerating any missing body part through the action of pluripotent stem cells called neoblasts (Reddien, 2018). Neoblasts maintain all cell types of the adult body through a process of constitutive cell turnover. Because planarians have >125 different adult cell types, these stem cells must be capable of choosing among a large array of possible cell fates (Fincher et al., 2018; Plass et al., 2018; Zeng et al., 2018; King et al., 2024). Fate specification can occur in neoblasts through the activation of transcription factors called fate-specific transcription factors (FSTFs), producing specialized neoblasts (Reddien, 2022). Specialized neoblasts divide and can produce daughter cells that act as migratory precursors (post-mitotic progenitors) for differentiated cell types (Eisenhoffer et al., 2008; Wenemoser and Reddien, 2010; Guedelhoefer and Sánchez Alvarado, 2012; van Wolfswinkel et al., 2014; Abnave et al., 2017; Reddien, 2022).

Fate choice in neoblasts can be regulated by position. For instance, eye-specialized neoblasts are formed in roughly the anterior third of the animal (Lapan and Reddien, 2011, 2012). However, this spatial regulation of stem cell fate specification is coarse, when compared to the precise positions of differentiated cells associated with complex planarian tissue architecture (Lapan and Reddien, 2011, 2012; Adler et al., 2014; Scimone et al., 2014a; van Wolfswinkel et al., 2014; Park et al., 2023). Because specialized neoblasts are produced in broad regions, they are often found far from their target tissue. As a result, individual specialized neoblasts are frequently found closer to other types of differentiated cell types than their target tissue itself (Park et al., 2023). Furthermore, neoblasts are spatially specified in a highly intermingled manner, in heterogeneous neoblast neighborhoods (Park et al., 2023). For instance, a muscle-specialized neoblast could have a neural, intestinal, epidermal, protonephridial, or other specialized neoblast type as its nearest neoblast neighbor (Park et al., 2023). These observations suggest that the regulation of differentiation programs in post-mitotic migratory progenitors is a crucial aspect to patterning and tissue maintenance. Neoblasts, themselves, are not highly migratory under homeostatic conditions; however, the post-mitotic progenitor cells that they produce serve as precursors and can remain in an immature and sometimes migratory state for days until reaching their target tissue (Saló and Baguñà, 1985; Eisenhoffer et al., 2008; Wenemoser and Reddien, 2010; Wagner et al., 2011; Guedelhoefer and Sánchez Alvarado, 2012; van Wolfswinkel et al., 2014; Abnave et al., 2017; Park et al., 2023). These findings generate a model for pattern maintenance during tissue turnover and formation in planarian regeneration in which specialized neoblasts generate intermingled post-mitotic progenitor classes that migrate to precise locations for differentiation (Park et al., 2023).

From a messy state of progenitor formation, order must arise. We hypothesized that terminal differentiation of post-mitotic progenitors is regulated to occur at precise positions to prevent disordered differentiation along migratory trails. This understudied mechanism could in principle be a major element of pattern formation from migratory progenitors in regenerative contexts. Understanding how migratory progenitors know when and where to terminally differentiate into a mature stationary cell is a central problem for understanding how systems of migratory progenitors generate and maintain pattern. The regenerative biology of planarians presents the opportunity to uncover mechanisms underlying migratory progenitor targeting and differentiation regulation, which could apply to numerous developmental and regenerative contexts across the animal kingdom.

Results

map3k1 RNAi results in the ectopic posterior differentiation of eye cells along the AP axis

We sought genes involved in processes that regulate where fate-specified progenitors differentiate through RNAi studies. We found a gene (map3k1, dd_5198) encoding a MAP3K1-like signaling protein for which RNAi resulted in a striking phenotype involving dispersed ectopic eyes and single eye cells (Figure 1A). The unique nature of this patterning phenotype raised the possibility that map3k1 has some role in regulating progenitor differentiation during targeting. Planarian map3k1 encodes a member of the MAP3K signaling protein family. Aside from a characteristic kinase domain, eukaryotic MAP3K1 proteins, including the planarian ortholog, possess unique accessory domains compared to other MAP3Ks; these include a PHD-like RING finger, a SWIM-type RING finger, and a TOG-like domain. These domains have been implicated in non-canonical MAP kinase signaling cascades that enable MAP3K1 to act as a ubiquitin ligase and a scaffold protein (Lu et al., 2002; Xia et al., 2007) (Figure 1- figure supplement 1A).

map3k1 RNAi results in ectopic eyes and ectopic isolated eye cells.

A. Top row: live image of a control RNAi animal followed by three examples of map3k1 RNAi animals with disorganized eyes (white arrows) (n=156/200 >1 ectopic eye, after 4 weeks of RNAi). Bottom row: control RNAi FISH images followed by three map3k1 RNAi examples visualizing OC cells (RNA probe pool to catalase/tyrosinase/glut3); PRNs are visualized with either an RNA probe to opsin (PRN cell bodies) or an anti-Arrestin antibody (PRN cell bodies and projections). The left and middle map3k1 RNAi FISH examples show single ectopic PRNs (opsin, left; anti-Arrestin, middle) and OC cells (catalase/tyrosinase/glut3) scattered below and around the eyes after 3 weeks of RNAi. The far FISH example shows ectopic OC cells and PRNs (anti-Arrestin) after 6 weeks of RNAi. Dorsal up. Scale bars, 100μm B. FISH showing ectopic PRNs (anti-Arrestin) and OC cells (catalase/tyrosinase/glut3) in the trunk and tail region of a map3k1 RNAi animal after 5 weeks of RNAi (n=10/10 with at least one cell in the trunk) (see panel 1A for control RNAi). Scale bar, 100μm. Magnified panels 1 and 2 scale bars, 50μm. White arrows point to all ectopic cells. C. FISH showing the tail regions of control and map3k1 RNAi animals; single OC (catalase/tyrosinase/glut3) cells are observed in the tail after one week of map3k1 RNAi (n=12/18 with at least one cell in the tail). Scale bar, 200μm. D. Top graph depicts the number of ectopic PRNs (1 week: n=15; 2 weeks: n=9; 3-4 weeks: n=22). Control shown at 3-4 weeks (n=11). Bottom graph depicts the number of ectopic OC cells (1 week: n=16; 2 weeks: n=7; 3-4 weeks: n=17) per map3k1 RNAi animal along the AP axis over time. Control shown at 3-4 weeks (n=18). map3k1 RNAi resulted in higher ectopic cell numbers along the AP axis (p<0.0001; Poisson generalized linear mixed model) compared to the control condition for both PRNs and OC cells at 3-4 weeks. E. Schematic comparing previously identified eye-patterning RNAi phenotypes (after ndk, wnt5, slit, and notum RNAi) with the map3k1 RNAi phenotype. F. FISH examples of eye-specialized neoblasts (ovo+; smedwi-1+ cells) in tails of control and map3k1 RNAi animals at 3 weeks of RNAi. Scale bar, 20μm. Right graph shows no significant difference (p=0.181; Permutation test, 10,000 permutations; two-tailed) in the frequency of eye-specialized neoblasts in the tails of control (n=10) and map3k1 RNAi (n=10) animals. All images, dorsal up. Bottom left numbers indicate the number of animals exhibiting the shown phenotype out of the total number of animals observed.

The bilaterally symmetric anterior location of planarian eyes represents the normal targeting location of eye progenitors (Lapan and Reddien, 2011, 2012). RNAi of map3k1 resulted in the gradual emergence of ectopic eyes posterior to the normal eye location with variance in their anterior-posterior (AP) and medial-lateral (ML) positions (Figure 1A). A similar map3k1 RNAi phenotype was independently described in (Lo and Petersen, 2025). Fluorescent in situ hybridization (FISH) experiments showed that both optic cup (OC) cells and photoreceptor neurons (PRNs) were present ectopically in clusters of cells and as individual, isolated cells (Figure 1A). Isolated, individual PRN or OC cell differentiation is rarely observed in the wild-type state. Although ectopic eye cells appeared most frequently in the normal zone of eye progenitor specification (the head), they also appeared throughout the trunk and tail – far from the canonical eye-progenitor specification zone (Figure 1B, C; Figure 1- figure supplement 1B). Isolated eye cells were observed in animal tails within 1-2 weeks of the first RNAi feeding when no, or very few other ectopic eye cells appeared between the head and tail (Figure 1C-D; Figure 1- figure supplement 1B). The far posterior ectopic eye cells that appeared in map3k1 RNAi animals were sparser than those in the head, and usually appeared as singletons rather than aggregates (Figure 1C; Figure1- figure supplement 1B).

The location in the planarian body where dispersed eye progenitors migrate towards, and target for differentiation, is referred to as the target zone (TZ) (Atabay et al., 2018). A variety of genes can be inhibited to result in eye formation in ectopic positions along the AP and ML axes, and are presumed to impact the location of the TZ: for instance, posterior TZ shifting following RNAi of nou darake (ndk) (Cebrià et al., 2002) and wntA (Kobayashi et al., 2007); anterior shifting following RNAi of notum (Hill and Petersen, 2015) and nr4A (Li et al., 2019); lateral shifting following RNAi of wnt5; and medial shifting following RNAi of slit (Oderberg et al., 2017; Atabay et al., 2018). However, in all of these cases organized ectopic eyes appear along particular AP or ML trajectories (Figure 1E, summary cartoon), consistent with alteration of the eye progenitor TZ along a single axis, but still with differentiation constrained to occur in a particular position on the orthogonal axis. By contrast, map3k1 RNAi ectopic eyes and eye cells were more disorganized, often being ectopic in both AP and ML position (Figure 1A). This distinct phenotype raised the possibility that map3k1 is required for some patterning process not previously disrupted by the inhibition of other patterning genes.

Patterning in planarians prominently involves constitutive and regional expression of genes constituting positional information, referred to as position control genes (PCGs) (Reddien, 2018). PCGs are predominantly expressed in planarian muscle (Witchley et al., 2013). ndk, notum, nr4A, wnt5, and slit are regulated in their spatial expression and are components of this PCG patterning system. Therefore, these genes likely influence progenitor-extrinsic cues that guide progenitors to particular locations. By contrast, map3k1 was not overtly expressed in a spatially restricted manner; instead, it was expressed broadly across tissues (Figure 1- figure supplement 1C).

The appearance of eye cells in the tail of map3k1 RNAi animals suggests that some eye-progenitor specification is likely occurring in the tails of these animals. Eye-specialized neoblasts can be recognized by the co-expression of the eye-specific FSTF ovo (Lapan and Reddien, 2012) and the neoblast marker smedwi-1 (Reddien et al., 2005). ovo+; smedwi-1+ neoblasts were observed in the midbody and tail of map3k1 RNAi animals (Figure 1F; Figure 1- figure supplement 1D, E). However, the frequency of observed ovo+ neoblasts in the tail was very low, and a similar low frequency of ovo+ neoblasts was observed in control animal tails (Figure 1F; Figure1- figure supplement 1E). There was therefore no overt change to the spatial pattern of eye-specialized neoblasts after map3k1 RNAi. This raises the possibility that there exists a natural low frequency of sporadic eye-neoblast specification events outside of the predominant, anterior eye progenitor specification zone and that these cells can ectopically differentiate into mature eye cells in the posterior when map3k1 is inhibited.

map3k1 prevents the posteriorization of some but not all anterior cell types

To assess whether the ectopic pattern of differentiated cells in map3k1 RNAi animals was specific to the eye, we visualized other differentiated tissue patterns. Laterally projecting neural branches in the central nervous system are normally restricted to the anterior planarian brain (Hyman, 1951). After map3k1 RNAi, however, ectopic brain branches labeled by the neuronal markers, choline acetyltransferase (chat) and prohormone convertase-2 (pc2), emerged from the two ventral nerve cords, appearing most frequently in the anterior and mid-body of the animal (Figure 2A; Figure 2- figure supplement 1A). In some instances, branches were observed in the tail (Figure 2- figure supplement 1B). Ectopic eyes sent axons (labeled with an anti-Arrestin antibody) that traveled along these ectopic neural branches and main nerve cord tracts (Figure 2A; Figure 2- figure supplement 1A,B). Ectopic brain branches that emerged from ventral nerve cords contained GluR+ (dd_16476) neurons, confirming that they are at least partly composed of normally brain-branch-restricted neurons (Figure 2B). GluR+ neurons are similarly found in ectopic brain branches observed after RNAi of ndk (Cebrià et al., 2002).

map3k1 RNAi results in posterior ectopic differentiation of some neurons and gland cells.

A. map3k1 RNAi animals exhibit posterior, ectopic brain branches that project dorsolaterally from ventral nerve cords (5 weeks of RNAi). These branches contain chat+ and pc2+ neurons. Photoreceptor axons (visualized with an anti-Arrestin antibody) extend along the ventral nerve cords and into ectopic brain branches. See also Figure 2- figure supplement 1A. Ventral up. Scale bar, 100μm. B. FISH showing gluR+ (dd_16476+) neurons in ectopic brain branches after 3 weeks of map3k1 RNAi. Scale bars, 200μm; higher magnification scale bar, 100μm. A, B. Ventral up. C. FISH images showing no change in cintillo+ neuron distribution in the head, but expansion of dd_17258+ neurons along the entire AP axis after 3 weeks of map3k1 RNAi. See also Figure 2-figure supplement 1B, C); Ventral up. Scale bars, 200μm. D. Graph showing no significant difference in the number of dd_17258+ neurons in the heads (AP_1) of map3k1 and control RNAi animals (p=0.410; Mann-Whitney test) but a significant difference in the number of ectopic cells observed along the entire AP axis (AP_2→AP_6: ***p<0.0006; Multiple Mann-Whitney tests). Counted animals underwent 3-4 weeks of RNAi; 2 replicates were used. No ectopic cintillo+ neurons were observed for both control and map3k1 RNAi animals. E. FISH images of unaffected (dd_9223) and affected (dd_7131 and dd_8476) parenchymal cell types in map3k1 RNAi animals (3 weeks of RNAi). Ventral up. Scale bar, 200μm. Right graph shows more dd_7131+ (p<0.0001; Negative binomial regression) and dd_8476+ (p<0.0001; Negative binomial regression) cells in the tails of map3k1 RNAi animals compared to control RNAi animals. Counted animals underwent 3-4 weeks of RNAi; 3 replicates each. Bottom left numbers indicate the number of animals with the result displayed in the image out of the total number of animals observed.

The anterior-restricted population of dd_17258+ neurons also displayed ectopic differentiation across the AP axis of map3k1 RNAi animals, extending to the tail tip (Figure 2C; Figure 2; figure supplement 1C, D). The correct number of dd_17258+ neurons, however, remained in the normal AP location (AP_1) (Figure 2D). By contrast, some other anterior neural populations, cintillo+ and glutamic acid decarboxylase+ (gad+) neurons, remained unaffected in map3k1 RNAi animals (Figure 2C; Figure 2- figure supplement 1C). These findings suggest that map3k1 is required for the normal AP restriction of a subset of neural cell types during tissue turnover.

To determine if tissue posteriorization in map3k1 RNAi occurred for non-neuronal cell types other than OC cells, we assayed gland cell populations (dd_9223, dd_7131+, and dd_8476+) that normally reside in the head, with a small fraction extending posteriorly to the pharynx and rarely to the tail. map3k1 RNAi animals showed posteriorization of gland cell distributions for the primarily anterior dd_7131+ and dd_8476+ populations (Figure 2E). As was the case with neuron types, not all gland cell types were affected by map3k1 RNAi. dd_9223+ cells, the most anteriorly restricted of the three gland cell populations assessed, did not change in distribution (Figure 2E). These data indicate that map3k1 broadly affects the spatial distribution of various differentiated cells during tissue maintenance, but that this role is restricted to a subset of cell types.

map3k1 inhibition causes ectopic anterior differentiation of pharynx progenitors

Another regional tissue that is maintained through turnover from regional progenitors is the planarian pharynx. Neoblasts that produce pharynx progenitors are broadly located in the trunk region of the animal and express FoxA (Adler et al., 2014; Scimone et al., 2014a). Pharynx progenitors enter the pharynx through a connection to the body at the anterior end of the organ, and this requires that progenitors are capable of moving in multiple directions as a response to extrinsic cues. After 3 weeks of map3k1 RNAi ectopic single vitrin+ cells were present around the pharynx and at the anterior end of the typical FoxA+ zone (n=20/20 animals), even reaching the head region in some cases (Figure 3A). These ectopic cells occupied variable AP and ML locations between the original pharynx and the brain (Figure 3A, Figure 3- figure supplement 1A). This variable placement of ectopic pharyngeal cells on the ML axis was reminiscent of the patterning defect observed for eyes following map3k1 RNAi, described above. Scattered ectopic foci of cells expressing mhc-1 (a gene expressed in pharyngeal muscle) were present in the anterior half of the animal, and frequently near ectopic vitrin+ pharyngeal cells after 3 weeks of RNAi (Figure 3B). Clusters of pharyngeal cell types were present between the cephalic ganglia and in the pre-pharyngeal region by 6-8 weeks of map3k1 RNAi (Figure 3- figure supplement 1B).

map3k1 RNAi results in pharynx cell types in ectopic anterior locations.

A. FISH images of control and map3k1 RNAi animals showing anterior expansion and dispersal of vitrin+ (pharynx) single cells (white arrows) and clusters of cells (pink arrows and boxes) at variable positions along the AP and ML axes at 3 weeks (n=20/20; 2 replicates) and 4 weeks (n=12/12; 1 replicate) of RNAi, between the cephalic ganglia (1), near the ventral nerve cords (2), and lateral to the pharynx (lower right panel). Scale bars, 200μm; magnified images scale bars, 20μm. Control animals, 3 weeks of RNAi. B Left panels: FISH images showing anterior clusters of mhc-1+ (pharynx muscle) cells (pink arrows) in map3k1 RNAi animals. Middle panel shows ectopic mhc-1+cells and vitrin+ cells near the brain in map3k1 RNAi animals (n=5/8; 1 replicate, white arrows). Left and middle panels, 3 weeks RNAi. Bottom right panel shows ectopic large clusters of mhc-1+ cells between the cephalic ganglia after 6 weeks of RNAi (n=6/6; 1 replicate, white arrows). Ventral up. Scale bar, 100μm. C. FISH images of map3k1 RNAi animals showing NB.22.1e+ mouth cells anterior to the normal location (pink brackets; 3 weeks RNAi) and dispersed around the pharynx (white arrows; 8 weeks RNAi). Scale bar, 100μm. D. FISH images showing clusters of dd_554+ cells (white arrows) (intermediate pharynx progenitor population (Zhu et al., 2015)) and single dd_554+ cells dispersed around and anterior to the pharynx (vitrin) in 3-week map3k1 RNAi animals. Counted animals for all panels underwent 3-4 weeks of RNAi; 3 replicates were used. Ventral up. Scale bar, 200μm. E. FISH showing examples of FoxA+; smedwi-1+ cells in the head region of both control and map3k1 RNAi animals (white boxes). 3 weeks RNAi, 2 replicates. See also Figure 3 – Figure supplement 1E. Scale bars, 200μm; magnified image scale bars, 10μm. A, B, C, D. All panels, ventral up. Numbers in each panel indicate number of animals displaying the result shown in the image out of total animals observed.

The planarian mouth is an epidermal opening at the posterior end of the pharynx. Following 3 weeks of map3k1 RNAi, ectopic mouth cells marked by NB.22.1e appeared anterior to the typical mouth location as an anterior streak stemming from the original mouth. Rarely (n=3/18 animals), an ectopic focus of NB.22.1e+ mouth cells was observed in the tail, posterior to the pharynx (Figure 3- figure supplement 1C). After 8 weeks of RNAi, ectopic scattered mouth cells were present lateral to the midline (Figure 3C). Because differentiated pharynx cells were observed outside of the canonical pharynx progenitor-specification zone, we considered the possibility that pharynx progenitor specification itself occurred in ectopic locations following map3k1 RNAi. dd_554 transcripts mark a post-mitotic pharynx progenitor population (Zhu et al., 2015). Ectopic dd_554+ cells were present anterior to the pharynx in map3k1 RNAi animals, including in foci (Figure 3D; Figure 3- figure supplement 1D). FoxA+; smedwi-1+ cells prominently, but not necessarily exclusively, include pharynx progenitors. These cells are predominantly centrally restricted on the AP axis to the trunk (Adler et al., 2014; Scimone et al., 2014a). We observed a small frequency of FoxA+; smedwi-1+ cells near the brain in both control and map3k1 RNAi animals (Figure 3E, Figure 3- figure supplement 1E-F). These findings are consistent with the possibility that these FoxA+; smedwi-1+ cells include pharynx progenitors, and that ectopic differentiation of pharynx progenitors can occur outside of the predominant pharynx progenitor specification zone at a low frequency.

Complete posterior duplication of the pharynx has previously been observed to occur following ptk7, ndl-3, and wntP-2 RNAi (Scimone et al., 2016; Hill and Petersen, 2018), anterior duplication has been observed after roboA RNAi (Cebrià et al., 2007), and lateral duplication has been observed after wnt5 RNAi (Gurley et al., 2010). The map3k1 RNAi phenotype differs from these other patterning phenotypes, in that it involves greater disorganization, with the appearance of small clusters of pharyngeal cells and even single pharyngeal cells instead of only well-organized but ectopically placed pharynges. This scenario bears similarity to the phenotype for the eye: disorganized ectopic tissue differentiation including in small clusters and single cells. These findings are consistent with the possibility that map3k1 RNAi disrupts the regulation of progenitor targeting that normally results in differentiation being restricted to occur in precise locations.

The targeting and maintenance of tissues after map3k1 RNAi

Post-mitotic progenitors normally migrate to, and differentiate at, their target zone (TZ). Prior work indicates that in addition to the TZ, the target tissue itself can incorporate its fate-specified progenitors and promote progenitor differentiation, even at an ectopic location (Atabay et al., 2018; Hill and Petersen, 2018). For instance, surgically transplanting an ectopic eye outside of the TZ, but within the broad distribution of eye progenitors, results in a stable ectopic eye that incorporates progenitors to replace dying cells as part of turnover (Atabay et al., 2018). This is enabled by the fact that progenitors in wild-type animals are specified in broad regions, giving ectopic differentiated tissues access to a constant supply of progenitors (Lapan and Reddien, 2011, 2012). Thus, at least two system components appear to be capable of promoting progenitor differentiation: the TZ and the target tissue. The pharynx shows similar properties to the eye – with an ectopic pharynx being maintained through progenitor incorporation and differentiation (Hill and Petersen, 2018). An ectopic organ, however, will not regenerate upon its removal. Because the original TZ location of the tissue is unchanged in this situation, progenitors will target the correct location after resection of an ectopic organ (Atabay et al., 2018; Hill and Petersen, 2018).

Given the above reasoning, ectopic eyes in map3k1 RNAi animals could in principle be explained by TZ movement or expansion. To assess whether TZ alteration occurred following map3k1 RNAi, or some other explanation for the map3k1 RNAi phenotype is more likely, we resected all visible eyes in map3k1 RNAi animals to determine the location of new progenitor differentiation in the absence of a target tissue. Eye-resected map3k1 RNAi animals regenerated eyes at a comparable rate to control animals, and in the normal TZ location (Figure 4A; Figure 4-figure supplement 1A). Notably, however, map3k1 RNAi animals did not regenerate any resected ectopic eyes (Figure 4A). This indicates that the TZ location is maintained after map3k1 RNAi, and that at least some progenitors are capable of reaching the normal TZ location in the absence of an eye. These data are consistent with a model in which map3k1 does not primarily control the positional information read by progenitors, but instead affects the ability of progenitors to differentiate at proper locations in response to a normal positional information system – a possibility explored further below.

map3k1 RNAi animals display tissue-specific regeneration at target zones.

A. Top schematic depicts eye resection experimental design. Live images of control and map3k1 RNAi animals prior to (d0; left column) and the day after (d1; middle column) eye resection. Right column shows live and FISH images of control and map3k1 RNAi animals 10 days following eye resection. PRNs are visualized with anti-Arrestin (PRN cell bodies and projections) and an RNA probe to opsin (PRN cell bodies). Optic cup cells are visualized using a pool of catalase1/ tyrosinase/glut3 RNA probes. 3 weeks of RNAi was performed prior to resections; FISH and live images are from different animals. Dorsal, up. Scale bars, 200μm. See Figure 4 - figure supplement 1A for d0 FISH. B. Diagram of pharynx resection on the left. Live images at d1 and d10 after pharynx resection showing pharynx regeneration in the correct location for both map3k1 RNAi and control animals. Pharynx regeneration in map3k1 RNAi animals is disorganized; FISH contains RNA probes to vitrin (pharynx-specific), NB.22.1e (mouth and esophagus), and mhc-1 (pharynx muscle). See also Figure 4- figure supplement 1B for d0 FISH. Ventral up. Live images, scale bars, 200μm; FISH images, scale bars, 100μm. C. Live image of d10 head, trunk, and regenerating tail fragments. map3k1 RNAi tail fragments regenerate pharynges (white arrow) in a more anterior location compared to control animals. C, D. 3 weeks of RNAi was performed prior to resection. Scale bar, 100μm. D. FISH images showing anterior vitrin+ pharynx cells (white arrows) and NB.22.1e+ mouth cells (white brackets) in map3k1 RNAi day 10 trunk regenerates. All map3k1 RNAi trunks and tails fully regenerate eyes (anti-Arrestin). Day 10 tails regenerate pharynges in a more anterior location compared to control animals. Ventral up. Scale bar, 200μm. E. Top graph shows no difference in the AP location of original pharynges in control and map3k1 RNAi d10 trunk regenerates (p=0.1054; Welch’s t-test) but a significant difference between control original pharynges and ectopic pharyngeal cell clusters map3k1 RNAi trunks (****p<0.0001; Mann-Whitney test). Bottom graph shows an anterior shift in pharynx regeneration in map3k1 RNAi tail fragments compared to control animals (p<0.0001; Welch’s t-test) Numbers in each panel indicate number of animals displaying the result shown in the image out of total animals observed.

To determine if the pharynx target zone is also maintained in its normal, wild-type location in map3k1 RNAi animals, we removed the entire pharynx of map3k1 RNAi and control animals (Figure 4B, Figure 4- figure supplement 1B). By 10 days after pharynx resection, animals regenerated a pharynx in the original position, but with some disorganization and multiple pharyngeal structures forming in some cases (Figure 4B, Figure 4- figure supplement 1B, C). Thus, the case for the pharynx is more complex than for the eye. Regardless, these findings for the pharynx suggest that at least the normal central TZ remains and is not simply shifted anteriorly.

The location of de novo organ regeneration after map3k1 RNAi

To determine the location of progenitor targeting and de novo organ formation during regeneration, we amputated map3k1 RNAi animals into head, trunk, and tail fragments and analyzed them after 10 days of regeneration (Figure 4C). All fragments regenerated organs that they did not already contain at the time of amputation (Figure 4D). Two eyes formed in approximately normal locations in map3k1 RNAi head blastemas, rather than being posteriorly shifted or appearing in multiple locations initially. This suggests, like the results from the eye-resection experiments described above, that the TZ is regenerated in roughly the wild-type location in map3k1 RNAi animals.

Prior work in the planarian species D. japonica showed that map3k1 RNAi results in tail fragments regenerating pharynges in an anterior-shifted location (Hosoda et al., 2018). Consistent with this previously reported effect, pharynges regenerated more anteriorly in map3k1 RNAi S. mediterranea tail fragments, just posterior to the regenerating brain (Figure 4C-E). Notably, trunks also regenerated secondary pharynx-like aggregates very close to the brain that were underdeveloped and that appeared to interfere with structures around them (Figure 4D). Head fragments regenerated multiple pharyngeal structures (Figure 4- figure supplement 1D). Therefore, the situation for the pharynx in regeneration is more complex than that of the eye, similar to the findings for organ resections described above.

PCG expression domains are largely unaffected by map3k1 RNAi

As noted above, numerous PCGs can be inhibited to cause organ duplications. However, the map3k1 RNAi phenotype described so far is largely consistent with alteration of progenitor targeting behavior rather than global shifting of positional information. To directly examine the spatial maintenance of positional information in map3k1 RNAi animals, we labeled these animals with RNA probes for multiple PCGs. Anterior, posterior, and medial PCG expression domains were largely unaffected in map3k1 RNAi animals – including expression domains for sFRP-1, ndl-4, ndl-5, ndl-2, ndl-3, wntP-2, axinB, sp5, ptk7, wnt11-1, and slit (Figure 5A). The spatial distribution of axinB transcription – a read-out of the posterior-to-anterior Wnt activity gradient in planarians – was similar in control and map3k1 RNAi animals, indicating that Wnt activity was maintained regionally on the AP axis (Figure 5A) (Iglesias et al., 2011; Reuter et al., 2015; Stückemann et al., 2017; Tewari et al., 2019).

Positional information remains largely unaffected in map3k1 RNAi animals

A. FISH panel of position control gene expression (sFRP-1, ndl-4, ndl-5, ndl-2, ndl-3, wntP-2, axinB, sp5, ptk7, wnt11-1, and slit) shows no obvious changes to expression domains in map3k1 RNAi animals. Animals from 3 and 4 weeks of RNAi were used. Pink arrowheads mark the end of PCG expression domains. Scale bars, 200μm. Ventral up. B. FISH example showing some dispersion of notum+; chat+ cells in the brain of a map3k1 RNAi animal after 3 weeks of RNAi (sample numbers from 3 and 4 weeks of RNAi). No obvious changes in wnt1+ posterior pole organization was observed after 2 and 3 weeks of RNAi (example FISH, 2 weeks RNAi). Scale bars, 200μm. Ventral up. C. Top row panels show example FISH images of opsin+(PRNs), dd_17258+ (neuron type), and vitrin+(pharynx) cells outside of typical PCG expression domains (ndl-5, ndl-5, and wntP-2, respectively). Bottom row panels show example FISH images of opsin+, dd_17258+, and vitrin+ cells inside atypical PCG expression domains (wntP-2, wnt11-1, and ndl-5/ndl-2, respectively). Far right panels on top and bottom, ventral up. All other panels, dorsal up. Animals from 3 and 4 weeks of RNAi were used. Scale bars, 200μm. Numbers in each panel indicate number of animals displaying the result shown in the image out of total animals observed.

The notum+ anterior pole, which is produced from neoblast-derived and migratory FoxD+ progenitors (Roberts-Galbraith and Newmark, 2013; Scimone et al., 2014b; Vogg et al., 2014), was present, but displayed some dispersal of cells after map3k1 RNAi (Figure 5B; Figure 5- figure supplement 1A). notum+; chat+ brain cells in the cephalic ganglia also appeared more dispersed in map3k1 RNAi animals, and this pattern was somewhat reminiscent of the eye and pharynx phenotypes. notum+ cell disorganization was also exaggerated during tail regeneration, when the animal is challenged with generating a new head (Figure 5- figure supplement 1B). Despite some anterior pole dispersal, the posterior pole appeared mostly normal and regenerating tails were still able to obtain proper PCG expression domains by day 4 (Figure 5B; Figure 5- figure supplement 1C).

Ectopic tissues in map3k1 RNAi animals formed outside of the PCG expression domains that they are normally restricted to; for instance, opsin+ and dd_17258+ neurons formed outside of the ndl-5+ domain and vitrin+pharyngeal cells formed outside of the wntP-2+ domain (Figure 5C). Ectopic differentiated cells were also observed inside of PCG expression domains that they are usually not found within; for instance, opsin+ and dd_17258+ neurons formed within a domain expressing wntP-2 and wnt11-1, and vitrin+ cells formed within ndl-2 and ndl-5 expression domains (Figure 5C). In map3k1 RNAi tail fragments, pharynges regenerated partially outside of the normal wntP-2+ expression domain, despite wntP-2 having a known role in establishing trunk identity (Figure 5- figure supplement 1D). These findings are consistent with the results described above: eyes are maintained and regenerated at the normal TZ, indicating that the defining positional information for the TZ remains at approximately the correct location in map3k1 RNAi animals. The fact that ectopic eyes emerge over time in a disordered fashion suggests that it is progenitor targeting for differentiation at precise locations that is affected by map3k1 RNAi rather than positional information itself.

map3k1 is expressed in neoblasts and migratory post-mitotic progenitors

Recent scRNA-seq work has annotated fate-associated clusters of planarian neoblasts and post-mitotic progenitors (King et al., 2024). If map3k1 acts in progenitors to regulate their differentiation, it should be transcribed in these cells. Indeed, map3k1 transcripts were present broadly across neoblast and post-mitotic progenitor clusters (Figure 5- figure supplement 2A, B), including for the eye, neural classes affected by map3k1 RNAi and the pharynx (Figure 5- figure supplement 2C-I). Whereas expression data alone do not necessarily indicate the site of action of map3k1, the data are consistent with the possibility that map3k1 can act within migratory progenitors.

map3k1 is required for restricting differentiation of eye progenitors along their migratory path to the target zone

Eye and pharynx progenitors are normally specified in restricted domains (Lapan and Reddien, 2011, 2012; Adler et al., 2014; Scimone et al., 2014a; Atabay et al., 2018), referred to here as progenitor specification zones. Within these domains, progenitors can migrate to reach their target tissue at their TZ, where they differentiate (Atabay et al., 2018; Hill and Petersen, 2018). We sought to test the possibility that ectopic cell differentiation, in map3k1 RNAi animals, is a result of premature progenitor differentiation at positions along the normal migratory path before reaching the TZ. An alternative scenario we considered is that ectopic progenitor specification at some distant location occurs after map3k1 RNAi and requires excessively long-range progenitor migration to reach the TZ, ultimately resulting in ectopic differentiation.

Neoblasts can be killed with irradiation (Bardeen and Baetjer, 1904), and lead shielding can be utilized in X-irradiation experiments to locally protect neoblasts, resulting in neoblasts being present only in a restricted field (Dubois, 1948; Guedelhoefer and Sánchez Alvarado, 2012; Abnave et al., 2017). Neoblasts expand slowly from these restricted regions (barring amputation) (Saló and Baguñà, 1985), but their postmitotic descendant cells serve as precursors that can readily migrate to target tissues (Guedelhoefer and Sánchez Alvarado, 2012; Abnave et al., 2017; Park et al., 2023). We utilized lead shielding over the top half of the animal to preserve the region of eye neoblasts, the place where eye progenitors are normally born, from X-irradiation (Figure 6A). 48 hours after irradiation, we initiated RNAi of map3k1 to observe the behavior of eye progenitors – restricted to be born within their normal, primary eye progenitor specification zone (the zone of surviving neoblasts). 12-14 days after the initiation of map3k1 RNAi in these partially irradiated animals, ectopic eye cells and ectopic dd_17258+ neurons were apparent (Figure 6A, Figure 6- figure supplement 1A, B). This indicates that ectopic differentiation can occur after map3k1 RNAi even from neoblasts restricted to undergo fate specification in the normal location.

map3k1 RNAi eye progenitors prematurely differentiate along a normal migratory path

A. Schematic of head-shielded irradiation experimental design. Animals were fixed 12-14 days following the first map3k1 RNAi feeding (2 days after irradiation), depending on health by inspection. Right graph shows ectopic differentiation events were more likely to occur in the anterior half of the animal (surviving neoblast region; marked by a smedwi-1+ RNA probe) versus outside of (distal or proximal to) the neoblast region for both opsin+and dd_17258+ neurons (p<0.0001; Binomial exact test). All observed ectopic PRNs (opsin) and dd_17258+ neurons outside of the neoblast region were present proximal to (within 50μm) the neoblast region. Bottom FISH panels depict examples of ectopic PRN (anti-Arrestin) and dd_17258+ neuron differentiation events within the neoblast region. Scale bar, 200μm; magnified boxes 1-4 scale bars, 50μm. B. Schematic of tail-shielded irradiation experimental design. Animals were fixed 10-12 days following the first map3k1 RNAi feeding (2 days after irradiation), depending on health by inspection. Right graph shows ectopic events were more likely to occur in the tail (area of surviving neoblasts) versus outside of (distal or proximal to) the neoblast region for dd_17258+ neurons (p<0.0001; Binomial exact test). All observed ectopic PRNs dd_17258+ neurons outside of the neoblast region were present proximal to (within 50μm) the neoblast region. FISH panels depict examples (white arrows) of ectopic PRNs (opsin) and dd_17258+ neurons in the tail of a tail-shielded, irradiated map3k1 RNAi animal. Scale bars, 200μm; magnified boxes 1-5 scale bars, 50μm. C. Schematic of EdU-labeled graft transplant experimental design; bottom graph shows a significant number of total ectopic eye cells (****p<0.0001; Mann-Whitney test) and ectopic EdU-positive eye cells (**p=0.002; Mann-Whitney test) in recipient wild-type animals compared to control. Animals after 2 and 3 weeks of RNAi prior to the EdU-pulse were used for transplantation. D. FISH example of EdU-positive ectopic eye cells (white arrows) differentiated in wild-type animals (n=13/20) with EdU-positive ectopic cells; n=19/20 exhibited any ectopic eye cells outside of the transplant area). Scale bars, 200μm; zoom in scale bars, 20μm. A, B, C, D. All panels are dorsal up. Numbers in each panel indicate number of animals displaying the result shown in the image out of total animals observed.

We next sought to confirm that the appearance of ectopic cells required the local production of new progenitors from neoblasts. We utilized lead shielding over the posterior half of wild-type animals and performed X-irradiation to preserve neoblasts only in the tail region (Figure 6B). We then performed map3k1 RNAi over a 10-12-day period and assessed animals for ectopic cell differentiation. The AP location of ectopic cells in these animals was most frequently in the area of remaining neoblasts (labeled with a probe to smedwi-1 transcript), with a small frequency of ectopic differentiation events occurring outside of this area, but still proximal to it (Figure 6B).

To assess the behavior of map3k1 RNAi eye progenitors in the normal eye progenitor specification zone and in a non-RNAi host environment, we labeled neoblasts with EdU after 2-3 weeks of map3k1 RNAi, then transplanted a small pre-pharyngeal tissue fragment from these animals into the pre-pharyngeal region of unlabeled, wild-type host animals (hosts had not experienced map3k1 RNAi) (Figure 6C). Transplant recipient animals were then fixed 12 days later. EdU-positive cells migrated out of the transplantation region and differentiated into eye cells at both the normal eye location (at the TZ) and at ectopic locations in the host environment (Figure 6C, D; Figure 6- figure supplement 1C). Additional ectopic eye cells were observed following transplantation that were not EdU-positive, likely because of incomplete labelling or because they were born multiple divisions after the EdU pulse was delivered (Figure 6- figure supplement 1D). These observations indicate that eye progenitors originating from the normal specification zone can erroneously differentiate before reaching their target location, including in a non-RNAi host environment.

Differentiated cells in the wrong organ of map3k1 RNAi animals

We administered map3k1 dsRNA feedings for RNAi and let animals undergo normal long-term tissue turnover to observe the consequences of errors in progenitor targeting on tissue pattern. Eight weeks of map3k1 RNAi led to differentiated pharynx muscle (mhc-1) cells within the cephalic ganglia (Figure 7A; Figure 7- figure supplement 1A). Single epidermal (NB.22.1e) cells were also observed, at a low frequency (n=4/18), within eyes after eight weeks of map3k1 RNAi. Tissue incorporation errors did not require many RNAi feedings to manifest. After only 3 weeks of map3k1 RNAi and tail amputation, day 10 tail regenerates (n=9/24) displayed ectopic vitrin+ pharyngeal cells in eyes (Figure 7B; Figure 7- figure supplement 1B). These striking cell-organization defects were not previously observed in the patterning phenotypes of PCG RNAi animals. We suggest that this defect highlights the risk to tissue architecture of not enacting tight spatial regulation of differentiation, especially in a biological context where progenitors are spatially dispersed and migratory. Ectopic differentiated cells can become incorporated into inappropriate tissue environments where they would normally never be observed, potentially disrupting tissue structure and function.

map3k1 RNAi results in differentiation in incorrect organs and teratoma formation

A. FISH showing clusters of mhc-1+ cells within the lobes of the brain and ventral nerve cords, and NB.22.1e+cells (epidermis, mouth) present within the eye after 8 weeks of map3k1 RNAi. Brain, ventral up; eye, dorsal up. Scale bars, 100μm. B. FISH examples of vitrin+ (pharynx) cells present within eyes (anti-Arrestin) at day 10 tail regeneration, following 3 weeks of map3k1 RNAi. See also figure 7- figure supplement 1B. Dorsal up. Scale bars, 200μm; scale bars for magnified eye images, 50μm. C. Left panels: live images of small growths in animals at 12 weeks of map3k1 RNAi feedings, accompanied by DAPI images of similarly positioned growths. Right panels: live images of advanced teratomas (pink arrows) in animals at 12 weeks of map3k1 RNAi, accompanied by a FISH example showing photoreceptors (opsin and anti-Arrestin) and OC cells (tyrosinase, catalase1, glut3) scattered in and around the teratomas. Bottom left animal, 8 weeks RNAi. C, D. Scale bars, 200μm. Dorsal up. D. Left panels: FISH images showing chat+, mhc-1+, and NB.22.1e+ cells are common in teratomas, often excluding vitrin+ (pharynx) and mag-1+ (gland cells). Scale bars, 100μm; panels 1-4 scale bars, 50μm. Right panels show examples of other cell types commonly found in outgrowths: cintillo+ (neuron), dd_17258+ (neuron), dd_3534+ (neuron), estrella+ (glia), anti-Arrestin+ (photoreceptors), NB.22.1e+, lamin+ (mouth and epidermis), and colF-2+ (muscle). 6-8 weeks of RNAi. Scale bars, 100μm. Dorsal up. Numbers in each panel indicate number of animals displaying the result shown in the image out of total animals observed.

map3k1 RNAi animals develop teratomas

All map3k1 RNAi animals ultimately developed tissue growths within 8-12 weeks, simply from undergoing tissue turnover without injury (Figure 7C; Figure 7- figure supplement 1C). These growths predominantly formed in the anterior of the animal and presented as a heterogeneous collection of cell and tissue types, and consistently contained clusters of neurons and muscle cells (Figure 7D; Figure 7- figure supplement 1D). Pharynx (vitrin+) cells and gland cells (mag-1+) were not as common in lateral outgrowths (Figure 7D). Cell types that were present in these growths included eye cells, neurons from the central nervous system (cintillo+, dd_17258+, and dd_3524+ cells), glia, muscle cells, and epidermal cells (Figure 7D). We considered these aberrant growths with a heterogeneous collection of intermingled tissues to be teratomas. A similar teratoma formation defect has been observed in planarians with a defect in progenitor migration caused by integrin RNAi (Bonar and Petersen, 2017; Seebeck et al., 2017). It is known that differentiated tissues, such as the eye, can trap their own progenitors and lead to their differentiation (Atabay et al., 2018; Hill and Petersen, 2018). This suggests that ectopic differentiation in inappropriate locations in map3k1 RNAi animals can result in the trapping of additional progenitors and can ultimately lead to an inappropriate aggregate of differentiated cells with aberrant pattern and organization. This defect highlights a further risk to tissue architecture if targeting and differentiation of migratory progenitors is not tightly controlled.

Discussion

Planarians display continuous turnover of adult tissues through the fate specification and differentiation of adult stem cells called neoblasts (Reddien, 2018). Fate specification in neoblasts can occur regionally (such as in the head for eye neoblasts) but is still spatially broad and intermingled (Park et al., 2023). Fate-specified neoblasts (specialized neoblasts) produce progeny cells that serve as precursors, referred to here as post-mitotic progenitors. These post-mitotic progenitors migrate to target locations to produce highly patterned anatomy. We suggest that differentiation is restricted during migratory targeting as an essential component of pattern formation, with the map3k1 RNAi phenotype indicating the existence and purpose of this element of patterning (Figure 8). We further suggest that progenitor targeting for local differentiation requires regulation from stem-cell-extrinsic signals, in the form of regionally expressed genes in muscle that comprise adult planarian positional information. How positional information interfaces with neoblasts and post-mitotic progenitors at the molecular level to regulate migratory assortment and differentiation only at precise locations is a fundamental problem of planarian regeneration and progenitor differentiation regulation. We suggest a model in which map3k1 acts as a brake on differentiation in stem cell progeny that can be lifted when suitable differentiation cues are encountered, either in the form of positional information (a target zone) or from interaction with target tissues (Figure 8A). We suggest that map3k1 is not required for the spatial distribution of progenitor-extrinsic differentiation-promoting cues themselves, but for progenitors to be restricted from differentiating until these cues are received (Figure 8B). Independent work in (Lo and Petersen, 2025) showed that map3k1 perturbation leads to ectopic progenitor differentiation in S. mediterranea providing supporting evidence for the findings presented here.

Model: Map3K1 restricts migratory progenitor differentiation until the correct target is reached

A. Schematic showing the inhibition of differentiation of a migratory progenitor cell, via map3k1, until reaching its target tissue at the target zone. B. Migratory precursors can differentiate in the incorrect PCG expression locations following map3k1 RNAi. C. Patterning abnormalities that can occur without suitable restriction of differentiation in migratory progenitors, demonstrating the patterning properties yielded by this mechanism.

Several observations support this model. First, PCG expression patterns themselves were largely normal following map3k1 RNAi. In the planarian species D. japonica, map3k1 RNAi was reported to cause an anterior expansion of sp5 expression in regenerating tails, but we did not note overt sp5 expression change in S. mediterranea. Additionally, recent work reported a slight posterior expansion of ndl-5 expression following map3k1 RNAi (Lo and Petersen, 2025), but this shift was small in magnitude. Second, ectopic differentiated cells were present directly within PCG expression domains that they are normally restricted from in map3k1 RNAi animals. Third, ectopic differentiation after map3k1 RNAi was more spatially disorganized than is typically observed for patterning phenotypes that occur following the shifting of PCG expression domains. For example, a posterior line of ectopic eyes emerges in ndk RNAi animals (Cebrià et al., 2002) and a lateral line of ectopic eyes emerges in wnt5 RNAi animals (Atabay et al., 2018), whereas ectopic eye cells emerged across AP and ML axes in a disorganized manner in map3k1 RNAi animals. Fourth, ectopic cells following map3k1 RNAi were frequently isolated, as opposed to appearing exclusively in organized aggregates or in ectopically placed organs. Fifth, during regeneration or following eye resection in map3k1 RNAi animals, progenitors could still be targeted to the correct target zone, indicating that the eye target zone remained present after map3k1 RNAi. Finally, transplantation of EdU-labeled tissue grafts from map3k1 RNAi animals into the pre-pharyngeal region of control animals showed instances of ectopic differentiation in control host tissue.

Map kinase signaling cascades are highly conserved in the animal kingdom and are involved in diverse processes that regulate dynamic cellular behaviors via activation of Map kinases (e.g., ERK1/2, p38, JNK) that can act as transcriptional modifiers (Widmann et al., 1999; Suddason and Gallagher, 2015). Map kinase kinase kinases (MAP3Ks) are one of the first activated proteins in these cascades, often responding to receptor tyrosine kinase signaling at the cell membrane. In mammals, there are 24 characterized MAP3Ks: MAP3K1 through MAP3K21, B-Raf, C-Raf, and A Raf, which activate downstream MAP2K proteins through phosphorylation. Among the MAP3K proteins, MAP3K1 orthologs are the only MAP3K with a PHD domain, which enables a role in ubiquitination as well as kinase activity (Pham et al., 2013; Suddason and Gallagher, 2015). In mammals, MAP3K1 has been implicated in cell proliferation, differentiation, and cell death signaling (Suddason and Gallagher, 2015).

The molecular mechanism by which map3k1 mediates its effect on planarian progenitors is still unclear. One scenario is that planarian post-mitotic progenitors are tuned to respond to a particular ECM or signaling environment (such as a PCG-related environment) to generate a molecular change that inactivates MAP3K1 signaling, such as by disengaging an RTK signal. Another possibility is that the progenitor migratory process itself could engage the MAP3K1 signal, enabling signal cessation with arrival at a target location. MAP3K1 can localize to focal adhesion complexes (Christerson et al., 1999; Cuevas et al., 2003), and has been implicated in integrin-mediated ECM-internalization mechanisms (Martinez et al., 2024) and interactions with Rho GTPases (Zhang et al., 2005; Chen and Cobb, 2006). In principle, one of these mechanisms could be connected to the regulation of planarian MAP3K1. When MAP3K1 is active it could result in a transcriptional state that prevents full expression of differentiated factors required for maturation, tissue incorporation, and cessation of migration.

The defect in the spatial restriction of progenitor differentiation following map3k1 RNAi can lead to dramatic tissue-patterning defects, including the differentiation of ectopic isolated cells (e.g., isolated photoreceptor neurons), the emergence of ectopic organs, differentiated cells from one tissue type being present within an incorrect organ, and teratoma formation (Figure 8C). Furthermore, in the case of distant progenitor specification events, these cells can be prevented from ever differentiating with a map3k1-controlled mechanism, potentially accommodating noise in the spatial precision of stem cell fate specification by enabling the pruning of distant specified progenitors from the system. These tissue-patterning defects highlight the importance of spatially restricting the differentiation of post-mitotic progenitors for maintaining and regenerating adult pattern. Similar regulation might prove important during developmental contexts in many organisms involving cell migration and could be particularly important in adult regeneration where tissue scale can be large and adult progenitors, at least in some contexts, can be challenged to migrate large distances before differentiating.

Important mysteries remain regarding the specific ways in which map3k1 regulates the patterning of different tissues. For example, ectopic dd_17258+ neurons in map3k1 RNAi animals were only out of place along the AP axis, and not on the ML axis; whereas ectopic eye cells were commonly out of place on both axes. This suggests that the ML component of differentiation regulation for dd_17258+ neurons might not be map3k1-dependent. The differentiated patterns of some studied neural and gland cell populations were also unaffected in map3k1 RNAi for unknown reasons, consistent with the possibility that map3k1 is one of multiple mechanisms for regulating differentiation in pattern formation.

There is evidence for the role of various Map kinases (e.g., ERK, MEK, RAS, p38) in planarian regeneration, particularly in blastema formation and wound response programs (Tasaki et al., 2011a; Tasaki et al., 2011b; Owlarn et al., 2017; Wang et al., 2020). It was suggested in another planarian species, D. japonica, that map3k1 has a role in the scaling and patterning of the trunk and head regions of regenerating animals (Hosoda et al., 2018). Additionally, map3k1 has been implicated in germ-cell proliferation and terminal differentiation of stem cells in the parasitic flatworm E. multilocularis through JNK signaling (Stoll et al., 2021). In (Lo and Petersen, 2025), gene function studies implicated both p38 and JNK Map kinases in the process regulated by map3k1. It will be of interest to further dissect the molecular role of map3k1 in planarian progenitor differentiation, and to determine whether map3k1 orthologs have similar roles in regulating differentiation in other regenerative contexts. map3k1 gene function is not well characterized in most invertebrate systems, including Drosophila and C. elegans which have no identified map3k1 orthologs (Widmann et al., 1999). Planarians therefore present an attractive model for dissection of map3k1 function.

Adult patterning systems in some organisms rely on spatially coarse and imperfect progenitor specification systems, requiring the targeted migration of progenitors and local differentiation cues at target locations. This process involves progenitor transitions from spatially broad and disorganized, to local and highly patterned structures (Xiong et al., 2013; Park et al., 2023). We suggest that in certain tissue formation processes from dispersed progenitors, cells will be prevented from differentiation until suitable extrinsic cues have been detected or cellular interactions have occurred, and that this regulation will be fundamental to pattern formation. We suggest that map3k1 acts within planarian progenitors to mediate such spatial restriction on differentiation, and that this is critical for preventing mistargeting of differentiation to incorrect locations, and to prevent teratoma formation.

Materials and methods

Animal husbandry and surgery

Asexual S. mediterranea clonal strain CIW4 was used for all experiments. Animals were cultured in static 1x Montjuic water (1.6 mmol/l NaCl, 1.0 mmol/l CaCl2, 1.0 mmol/l MgSO4, 0.1 mmol/l MgCl2, 0.1 mmol/l KCl, and 1.2 mmol/l NaHCO3 prepared in Milli-Q water) at 20°C. Amputations were performed under cold conditions (∼4C) with a scalpel. Animals were fed homogenized beef liver weekly, with water changed biweekly. Animals were starved for approximately seven days before experiments.

Whole-mount fluorescent in situ hybridization (FISH)

Animal mucus was removed using 5% N-acetylcysteine in PBS; animals were then fixed with 4% formaldehyde in PBST for 20 minutes, with rocking. Animals were then washed with PBST, incubated in 1:1 PBST: methanol, then stored in 100% methanol at −20C until ready for bleaching. Animals were moved into mesh baskets in a 24-well plate where all remaining steps were carried out. Animals were placed on a light source to bleach for 1.5 hours in a bleaching solution (5% formamide, 0.5x SSC, and 1.2% hydrogen peroxide). After two PBST washes, animals were then treated with 5 mg/ml Proteinase K for 10 minutes, followed by 4% formaldehyde post-fixation in PBST.

RNA probes were transcribed using DIG, FITC, or DNP-modified nucleotides, allowing signal amplification with DIG, FITC, and DNP antibodies conjugated to an HRP. Probes were diluted in Hybe solution (1:800) (50% deionized formamide, 5x SSC, 1 mg/mL yeast RNA, 1% Tween-20, 5% dextran sulfate), and left to incubate overnight. The following days, we performed antibody incubations at 4C overnight using anti-DIG-POD (1:1500, Roche; 10% western blocking solution (Roche) anti-FITC-POD (1:2,000, Roche; 5% horse serum, 5% western blocking solution), and anti-DNP-HRP (1:100, Perkin-Elmer; blocking solution with 10% inactivated Horse Serum). Tyramide signal amplification involved incubating in rhodamine (1:1,000), fluorescein (1:1,500), or Cy5 (1:300) in borate buffer (0.1M boric acid, 2M NaCl, pH 8.5) containing 0.0003% hydrogen peroxide for 10 minutes. Samples were incubated in 1% sodium azide for two hours to inactivate the HRP. Blocking and antibody incubations then occurred for detection of the second probe. Animals were incubated overnight in 1mg/mL DAPI solution at 4C. Animals were mounted on coverslips in ProLong Gold Antifade Mountant (Thermo Fisher).

EdU labeling and detection

F-ara-EdU (Click Chemistry Tools) was diluted in Dimethyl sulfoxide (DMSO) to 200mg/ml, then diluted in static 1x Montjuic water to 1.25mg/mL. Animals were split into 10 animals per well in a 12-well plate, then soaked in 1.25mg/mL EdU solution for 20 hours following 1 week of starvation. EdU solution was replaced with 5mg/mL Instant Ocean Sea salt dissolved in Milli-Q water. Prior to probe hybridization in the in situ hybridization protocol, following proteinase K and 4% formaldehyde incubations, cells were incubated in a “click reaction”-1% 100 mM CuSO4, 0.1% 10 mM TAMRA-Azide-fluor 545 (Sigma-Aldrich), and 20% 50 mM ascorbic acid in PBS-for 30 minutes in the dark, proceeded by 6 PBST washes and continuation of the probe hybridization step.

RNA interference

dsRNA for RNAi was prepared using in vitro transcription with T7 polymerase (Promega). Template was generated with PCR; primers contained T7 promoters. RNA was resuspended in water and dsRNA annealed. C. elegans unc-22 dsRNA was used as the negative control for all RNAi experiments. 50uL of homogenized beef liver was mixed with 25uL of dsRNA prep and 3uL of a 1:1 mixture of MilliQ water and red food coloring. map3k1 RNAi experiment durations ranged from 1 to 16 weeks of dsRNA feedings for RNAi. Time courses were conducted using animals fed 1 week, 2 weeks, and 3-4 weeks of dsRNA. Animals used to study outgrowth phenotypes were fixed between 8-12 weeks of RNAi, based on teratoma severity. Animals fed dsRNA for 3 and 4 weeks showed similar phenotype severity and were often analyzed as one group. Animals were given food for 1 hour, twice a week, for the first 8 feedings (4 weeks); animals were then given food once a week, for 30 minutes, for all subsequent feedings to prevent rapid growth and fissioning (separation of the tail from the body).

EdU transplantation assays

EdU plug transplants were performed using map3k1 or control RNAi animals, 12 hours following a 20-hour EdU pulse, as the donor to a recipient wild-type animal. Donor animals were anesthetized with 0.2% chlorotone solution, followed by an incubation in Holfreter’s solution, then placed on an ice block covered in Whatman filter paper moistened with 1x Montjuic water to surgically manipulate with a clean scalpel. EdU-positive map3k1 RNAi donor animals and wild-type recipients both had a center portion of their pre-pharyngeal regions removed. The pre-pharyngeal donor graft from the map3k1 RNAi animal was placed in the EdU-negative wild-type recipient’s pre-pharyngeal region. Recipients were then gently covered with cigarette paper soaked in chilled Holtfreter’s solution and transferred to a small petri dish with enough Holtfreter’s solution to cover the bottom of the dish. Petri dishes were put at 10C for 20 hours; the following day, transplant recipients were gently recovered and put into 1x Montjuic water containing 0.1% gentamicin (Gibco) to heal. Water was changed every two days, and animals were fixed at day 12 post-transplant.

Shielded irradiation

Animals were irradiated using an X-Rad320, Precision X-Ray Irradiation chamber. For shielded irradiation experiments, animals were anesthetized with 0.2% chlorotone, then arranged on Whatman filter paper in a petri dish sitting on ice. Animals were oriented to have their anterior half covered by the lead shield placed over the petri dish. Samples were placed in the irradiation chamber and exposed to 3,000 Rad of unidirectional X-irradiation. Animals were rescued with 1x Montjuic water and stored in 1x Montjuic water with 0.1% gentamicin (Gibco) to recover. Water was changed every two days. RNAi experiments were carried out starting two days after irradiation exposure. Anterior half-shielded animals were fed dsRNA for a period of 12-14 days before fixation, and posterior half-shielded animals were fed dsRNA for a period of 10-12 days before fixation; these ranges were dependent on the health of the animals at the time of fixation. If animals started showing slight signs of health decline (e.g., small lesions, slight head regression, bloating), they were fixed on that day.

Regeneration assays

All tissue resections and amputations were performed by placing animals on wet filter paper on top of a cold block to minimize movements during surgeries. Animals were kept moist throughout all procedures with 1x Montjuic water. Eyes were resected using a small scalpel. Curved edges were created by lightly tapping the tip of the scalpel on a clean, hard surface. The curved edge was used to poke and scoop out the eye in a poke and pull motion. Pharynges were resected by puncturing a diamond shape around the pharynx with a small scalpel, then gently removing the pharynx tissue. Animals were placed in 1x Montjuic water with 0.1% gentamicin (Gibco) to recover and water was changed every two days.

Image analysis and statistical analysis

FISH images were analyzed using Fiji Software. The AP axis was binned into 6 regions according to anatomical landmarks: AP_1 (head tip →bottom of the brain), AP_2 (bottom of the brain → top of the pharynx), AP_3 (top of the pharynx → middle of the pharynx), AP_4 (middle of the pharynx → bottom of the pharynx), AP_5 (bottom of the pharynx → half way between the bottom of the pharynx and the tail tip), and AP_6 (half way between the bottom of the pharynx and the tail tip → tail tip). Each animal had one data point in each of the 6 AP bins. map3k1 and control RNAi ectopic PRN and OC cell counts at 3-4 weeks were analyzed by generating a Poisson generalized linear mixed model, using AP bin as a random covariate. Any tail or head cell count data sets showing overdispersion – gland cells (dd_7131 and dd_8476) and FoxA+; smedwi-1− cell counts between the brain – were analyzed with a negative binomial regression, correcting for overdispersion. EdU transplant counts were analyzed with a Mann-Whitney-U test. Two-tailed permutation tests, using 10,000 permutations per test, were carried out for ovo+ cell counts in the tail because of the low sample number and zero-inflated dataset. A binomial exact test was carried out on ectopic cells inside versus outside the smedwi-1+ zone after shielded irradiation and map3k1 RNAi. Prism software was used to carry out Student’s t-tests, Mann-Whitney U tests, and Binomial exact tests. R studio was used to compute negative binomial regression, Poisson regression, and permutation tests.